### 2022-10-10 theoretical-consensus/article/article.tex
init commit

%% bare_jrnl.tex
%% V1.4b
%% 2015/08/26
%% by Michael Shell
%% see http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.8b or later) with an IEEE
%% journal paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/pkg/ieeetran
%% and
%% http://www.ieee.org/

%%*************************************************************************
%% Legal Notice:
%% This code is offered as-is without any warranty either expressed or
%% implied; without even the implied warranty of MERCHANTABILITY or
%% FITNESS FOR A PARTICULAR PURPOSE! 
%% User assumes all risk.
%% In no event shall the IEEE or any contributor to this code be liable for
%% any damages or losses, including, but not limited to, incidental,
%% consequential, or any other damages, resulting from the use or misuse
%% of any information contained here.
%%
%% All comments are the opinions of their respective authors and are not
%% necessarily endorsed by the IEEE.
%%
%% This work is distributed under the LaTeX Project Public License (LPPL)
%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
%% distributed and modified. A copy of the LPPL, version 1.3, is included
%% in the base LaTeX documentation of all distributions of LaTeX released
%% 2003/12/01 or later.
%% Retain all contribution notices and credits.
%% ** Modified files should be clearly indicated as such, including  **
%% ** renaming them and changing author support contact information. **
%%*************************************************************************

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. The IEEE's font choices and paper sizes can   ***
% *** trigger bugs that do not appear when using other class files.       ***                          ***
% The testflow support page is at:
% http://www.michaelshell.org/tex/testflow/

\documentclass[journal]{IEEEtran}

\usepackage{cite}

%\usepackage[pdftex]{graphicx}

%\usepackage{amsmath}

%\usepackage{algorithmic}

%\usepackage{array}

%\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}

%\usepackage{fixltx2e}

%\usepackage{stfloats}

%\usepackage{dblfloatfix}

%\usepackage{url}

\usepackage{hyperref}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Line breaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{The Units of Permissionless Consensus: \\ Towards Mobile and Edge Computing}
%
%
% author names and IEEE memberships
% note positions of commas and non breaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

\author{Eduardo Ribas Brito \\ eduardo.ribas.brito@ut.ee}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
% \markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
% {Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
...
\end{abstract}

% Note that keywords are not normally used for peer review papers.
\begin{IEEEkeywords}
IEEE, IEEEtran, journal, \LaTeX, paper, template.
\end{IEEEkeywords}

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
% \IEEEpeerreviewmaketitle

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{L}{ong} has been the time when consensus started 
to be defined as a fundamental problem of distributed systems \cite{pease1980reaching, lamport2019byzantine, lamport1983weak}. 
Generally, consensus means reaching an agreement between multiple parties 
in the potential presence of faulty individuals. As per multi-agent systems, 
interacting over computer networks, consensus is thought to be the result 
of a coordination effort such that those parties agree on some value at a 
given moment. Achieving consensus implies that the system shall be reliable 
and fault-tolerant. However, the consensus problem has been limited by some 
assumptions on the networks. The well-known "secure Byzantine-Fault-Tolerant 
multiparty consensus systems" that have been designed over the years are usually 
meant to work only with a set of known participants, faulty or not \cite{castro1999practical}. The 
other side of the coin is the permissionless consensus challenge, consisting of 
achieving agreement in an environment where the participants are unknown and 
untrusted \cite{nakamoto2008bitcoin, buterin2014next}. Plus, there are other intrinsic particularities of 
this type of networks, for example, their openness, and the lack of any kind of central 
authority. This adds another layer of complexity to the problem, as the participants 
are not only unknown and untrusted but can also join or leave the network at any time, 
freely choosing if they want to participate in the consensus protocol or not. Nevertheless, 
the problem of permissionless consensus can still be seen as a special case of the more 
general consensus and can still be formalized in the same way. In this paper, we will 
focus on consensus in permissionless systems, especially in the context of blockchain 
networks. We will reason about its meaningfulness, ultimately by trying to identify the units that 
may underpin consensus. We will also discuss the current state-of-the-art, with a 
particular interest in the shift of the consensus layer of these distributed networks 
to mobile and edge computing environments, for which computationally expensive consensus 
algorithms are impractical and unfair.

\section{Related Work}

\subsection{Classical Consensus}

The establishment of a definition for the problem of reaching agreement in a distributed 
system was pioneered by Lamport et al. in \cite{pease1980reaching}. 
The authors defined consensus as the problem of agreeing on a single value among a set of processes, 
in the presence of faulty entities. The first consensus algorithms were designed for 
synchronous systems, where the communication between the processes is reliable, and the 
delay is bounded. However, these initial attempts failed to cover the different types of 
faulty behaviour. Along with the establishment of the famous Byzantine Generals Problem, 
the first solutions, not only for dealing with treason, but also for unreliable communication 
channels, or any other kind of arbitrary byzantine behaviour, were 
also proposed by Lamport in \cite{lamport2019byzantine,lamport1983weak}. The solution was a synchronous 
mechanism that used a set of leaders to reach consensus. Multiple practical implementations 
and optimizations to this solution have been proposed in the literature.

\subsection{Asynchronous Byzantine Consensus}

The first asynchronous consensus algorithm was later proposed by Castro and Liskov in \cite{castro1999practical}. 
And naturally, after that work, many other asynchronous consensus algorithms have appeared. 
However, all of them are based on the assumption that the number of faulty processes is less than 
a certain threshold. Additionally, the assumption of a known set of participants is also made. This is a very strong assumption, 
as it is not always possible to know the participants beforehand as they may, for example, participate anonymously, or dynamically.

\subsection{Permissionless Consensus}

The advancements of the internet 
more than potentiated the revolution and what we now call the permissionless consensus problem was 
finally born. Without forgetting the previous attempts, the first practical permissionless consensus algorithm was 
proposed by Nakamoto in \cite{nakamoto2008bitcoin}. It is a proof-of-work consensus protocol that resembles a "replicated 
state machine" where the independent participants reach agreement not only about transactional values, 
but also about their order. "Proof-of-work is essentially one-CPU-one-vote" and this is the novelty 
introduced by Bitcoin \cite{pass2016hybrid,pass2016hybrid2}. The focus shifted 
for decentralized systems and after proof-of-work many other consensus mechanisms have been proposed, 
based on different consensus units, like proof-of-stake, proof-of-space, proof-of-burn, etc. The chaotic 
diversity of new consensus protocols gave also room for endless reviews, overviews and comparisons \cite{survey-dist-consensus, BAMAKAN2020113385, 8400278, 8629877, 9376868, BOURAGA2021114384}. 
The authors of these surveys often put multiple dimensions into comparison, like fault tolerance, scalability, or energy consumption,
and, among those, some focused their efforts on mechanisms that may work in resource-constrained networks \cite{9104713, 8661654, 8168250, SALIMITARI2020100212}. 
This paper will try to identify the common conclusions from these comparisons, while looking into the state-of-the-art and novel approaches for running permissionless
consensus protocols in mobile and edge devices.

% Subsection text here.

% needed in second column of first page if using \IEEEpubid
%\IEEEpubidadjcol

% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captionsoff option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure}

% Note that the IEEE typically puts floats only at the top, even when this
% results in a large percentage of a column being occupied by floats.

% An example of a double column floating figure using two subfigures.
% (The subfig.sty package must be loaded for this to work.)
% The subfigure \label commands are set within each subfloat command,
% and the \label for the overall figure must come after \caption.
% \hfil is used as a separator to get equal spacing.
% Watch out that the combined width of all the subfigures on a 
% line do not exceed the text width or a line break will occur.
%
%\begin{figure*}[!t]
%\centering
%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
%\label{fig_second_case}}
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure*}
%
% Note that often IEEE papers with subfigures do not employ subfigure
% captions (using the optional argument to \subfloat[]), but instead will
% reference/describe all of them (a), (b), etc., within the main caption.
% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
% labels, the optional argument to \subfloat must be present. If a
% subcaption is not desired, just leave its contents blank,
% e.g., \subfloat[].

% An example of a floating table. Note that, for IEEE style tables, the
% \caption command should come BEFORE the table and, given that table
% captions serve much like titles, are usually capitalized except for words
% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
% and up, which are usually not capitalized unless they are the first or
% last word of the caption. Table text will default to \footnotesize as
% the IEEE normally uses this smaller font for tables.
% The \label must come after \caption as always.
%
%\begin{table}[!t]
%% increase table row spacing, adjust to taste
%\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
%\caption{An Example of a Table}
%\label{table_example}
%\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
%\begin{tabular}{|c||c|}
%\hline
%One & Two\\
%\hline
%Three & Four\\
%\hline
%\end{tabular}
%\end{table}

% Note that the IEEE does not put floats in the very first column
% - or typically anywhere on the first page for that matter. Also,
% in-text middle ("here") positioning is typically not used, but it
% is allowed and encouraged for Computer Society conferences (but
% not Computer Society journals). Most IEEE journals/conferences use
% top floats exclusively. 
% Note that, LaTeX2e, unlike IEEE journals/conferences, places
% footnotes above bottom floats. This can be corrected via the
% \fnbelowfloat command of the stfloats package.

% \section{Conclusion}
% The conclusion goes here.

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

% \appendices
% \section{Proof of the First Zonklar Equation}
% Appendix one text goes here.

% % you can choose not to have a title for an appendix
% % if you want by leaving the argument blank
% \section{}
% Appendix two text goes here.

% % use section* for acknowledgment
% \section*{Acknowledgment}

% The authors would like to thank...

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/

%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)

\bibliography{bib}{}
\bibliographystyle{unsrt}

% biography section
% 
% If you have an EPS/PDF photo (graphicx package needed) extra braces are
% needed around the contents of the optional argument to biography to prevent
% the LaTeX parser from getting confused when it sees the complicated
% \includegraphics command within an optional argument. (You could create
% your own custom macro containing the \includegraphics command to make things
% simpler here.)
%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
% or if you just want to reserve a space for a photo:

% \begin{IEEEbiography}{Michael Shell}
% Biography text here.
% \end{IEEEbiography}

% % if you will not have a photo at all:
% \begin{IEEEbiographynophoto}{John Doe}
% Biography text here.
% \end{IEEEbiographynophoto}

% insert where needed to balance the two columns on the last page with
% biographies
%\newpage

% \begin{IEEEbiographynophoto}{Jane Doe}
% Biography text here.
% \end{IEEEbiographynophoto}

% You can push biographies down or up by placing
% a \vfill before or after them. The appropriate
% use of \vfill depends on what kind of text is
% on the last page and whether or not the columns
% are being equalized.

%\vfill

% Can be used to pull up biographies so that the bottom of the last one
% is flush with the other column.
%\enlargethispage{-5in}

% that's all folks
\end{document}

### 2022-10-10 practical-consensus/main.go

@@ -1,0 +1,15 @@
+package main
+
+func main() {
+	peers := []string{}
+
+	for i := 0; i < 3; i++ {
+		peer, err := startNode(peers)
+		if err != nil {
+			panic(err)
+		}
+		peers = append(peers, peer)
+	}
+
+	select {}
+}

### 2022-10-10 practical-consensus/miner.go

@@ -1,0 +1,118 @@
+package main
+
+import (
+	"context"
+	"crypto/sha256"
+	"fmt"
+	"log"
+	"math/rand"
+	"strings"
+
+	"github.com/libp2p/go-libp2p"
+	"github.com/libp2p/go-libp2p/core/host"
+	"github.com/libp2p/go-libp2p/core/peer"
+	"github.com/libp2p/go-libp2p/core/peerstore"
+
+	ma "github.com/multiformats/go-multiaddr"
+)
+
+type Node struct {
+	Host       host.Host
+	Blockchain Blockchain
+}
+
+type Blockchain []Block
+
+var currentVal int
+
+const difficulty = 6
+
+type Block struct {
+	Val   int
+	Nonce int
+	Hash  string
+}
+
+func (b Block) String() string {
+	return fmt.Sprintf("Val: %d, Nonce: %d, Hash: %s", b.Val, b.Nonce, b.Hash)
+}
+
+func newPeer(peers []string) {
+
+	var logger = log.New(log.Writer(), "miner: ", log.Flags())
+
+	// create the miner node
+	node, err := createMinerNode(peers)
+	if err != nil {
+		logger.Printf("Error creating miner node: %v", err)
+		return
+	}
+
+	logger.SetPrefix(fmt.Sprintf("miner %v: ", node.Host.ID()))
+
+	logger.Println("info: miner node started")
+
+	// start mining
+	mine(node, logger)
+}
+
+func createMinerNode(peers []string) (Node, error) {
+	// start a libp2p node that listens on a random local TCP port
+	host, err := libp2p.New(
+		libp2p.ListenAddrStrings("/ip4/127.0.0.1/tcp/0"),
+	)
+
+	if err != nil {
+		return Node{}, fmt.Errorf("error: create node: %v", err)
+	}
+
+	node := Node{Host: host, Blockchain: Blockchain{}}
+
+	if len(peers) == 0 {
+		genesis := Block{Val: 0, Nonce: 0}
+		genesis.Hash = calculateHash(genesis)
+		node.Blockchain = append(node.Blockchain, genesis)
+	} else {
+		// TODO: connect to peers
+		for _, p := range peers {
+			paddr, err := ma.NewMultiaddr(p)
+			if err != nil {
+				return Node{}, fmt.Errorf("error: create peer address: %v", err)
+			}
+			info, err := peer.AddrInfoFromP2pAddr(paddr)
+			if err != nil {
[…]

### 2022-10-10 practical-consensus/peer.go

@@ -1,0 +1,157 @@
+package main
+
+import (
+	"context"
+	"encoding/json"
+	"fmt"
+	"log"
+	"math/rand"
+	"time"
+
+	"github.com/libp2p/go-libp2p"
+	"github.com/libp2p/go-libp2p/core/host"
+	"github.com/libp2p/go-libp2p/core/peer"
+	"github.com/libp2p/go-libp2p/core/peerstore"
+
+	pubsub "github.com/libp2p/go-libp2p-pubsub"
+	ma "github.com/multiformats/go-multiaddr"
+)
+
+func startNode(peers []string) (string, error) {
+	var logger = log.New(log.Writer(), "peer: ", log.Flags())
+
+	node, err := createNode(peers)
+	if err != nil {
+		logger.Printf("error: create node: %v", err)
+		return "", err
+	}
+
+	topic, sub, err := createPubSub(node, logger)
+	if err != nil {
+		logger.Printf("error: create pubsub: %v", err)
+		return "", err
+	}
+
+	logger.SetPrefix(fmt.Sprintf("peer %v: ", node.ID().ShortString()))
+
+	logger.Println("info: peer started")
+
+	go subscribe(node, sub, logger)
+
+	go publish(node, topic, logger)
+
+	return getPeerMultiAddress(node), nil
+}
+
+func createNode(peers []string) (host.Host, error) {
+	// start a libp2p node that listens on a random local TCP port
+	node, err := libp2p.New(
+		libp2p.ListenAddrStrings("/ip4/127.0.0.1/tcp/0"),
+	)
+	if err != nil {
+		return nil, err
+	}
+
+	// add the peers to the peerstore
+	for _, p := range peers {
+		paddr, err := ma.NewMultiaddr(p)
+		if err != nil {
+			return nil, fmt.Errorf("error: create peer address: %v", err)
+		}
+		info, err := peer.AddrInfoFromP2pAddr(paddr)
+		if err != nil {
+			return nil, fmt.Errorf("error: create peer info: %v", err)
+		}
+
+		node.Peerstore().AddAddr(info.ID, paddr, peerstore.PermanentAddrTTL)
+		if err := node.Connect(context.Background(), *info); err != nil {
+			return nil, fmt.Errorf("error: connect to peer: %v", err)
+		}
+	}
+
+	return node, nil
+}
+
[…]

### 2022-10-10 theoretical-consensus/article/article.tex
page 02

@@
 However, all of them are based on the assumption that the number of faulty processes is less than 
-a certain threshold. Additionally, the assumption of a known set of participants is also made. This is a very strong assumption, 
-as it is not always possible to know the participants beforehand as they may, for example, participate anonymously, or dynamically.
+a certain threshold. Additionally, the assumption of a known set of participants is also made, as well as their roles 
+in the consensus protocol. These are very strong assumptions that limit the challenges that can be addressed, 
+for example, the impossibility to know the participants beforehand as they may participate anonymously, or dynamically.
 
@@
 state machine" where the independent participants reach agreement not only about transactional values, 
-but also about their order. "Proof-of-work is essentially one-CPU-one-vote" and this is the novelty 
+but also about their order - naturality forming the underlying structure of what is now known as a blockchain. 
+"Proof-of-work is essentially one-CPU-one-vote" and this is the novelty 
 introduced by Bitcoin \cite{pass2016hybrid,pass2016hybrid2}. The focus shifted 
@@
 consensus protocols in mobile and edge devices.
+
+\section{The need for permissionless consensus}
+
+\subsection{Permissioned vs Permissionless}
+
+Following the line of the previous sections, it has been reasoned about the need 
+for permissionless consensus when there are already well known and established consensus protocols
+that work in trusted environments \cite{castro1999practical}. However, even those protocols have
+their own limitations, not only in terms of trust, fault-tolerance, centrality, permissioning,
+or bottlenecking, but also in terms of scalability \cite{miller2016honey}. Trying to put some effort on decentralization, 
+Byzantine-Fault-Tolerant consensus protocols are not known for their performance when the network grows
+in size. The more participants there are, the more messages need to be exchanged
+between them, and the more time it takes to reach consensus, even if assuring deterministic finality \cite{decker2016bitcoin}. 
+This is a problem that is not only related to the number of participants, but also to the
+communication fashion and bandwidth, and the computational capacity of the devices that
+participate in the consensus protocol. 
+Summing up, the need for permissionless consensus is then justified by the fact that
+permissioned protocols are not compatible with the requirements of the new generation of
+distributed systems, especially in the context of blockchain networks. 
+These requirements include dealing with today's sparse networks of anonymously and dynamically
+participating devices, without interrupting consensus and while generally battling Sybil attacks \cite{survey-dist-consensus}.
+Fundamentally, the permissionless consensus problem is the need for a consensus protocol that
+can be run in a "zero-trust" environment, where the participants are unknown and untrusted,
+and where the network is sparse and unpredictably unreliable.
+
+\subsection{Allowances and goals}
+
+Technically, permissionless environments allow for larger networks
+that depit lower connectivity between the participants. Operationally,
+everything is expected to happen in an asynchronous or partially synchronous fashion, 
+and the number of transactions is expected to be smaller than the 
+permissioned counterparts. Nevertheless, participation is free, and the
+governance is not centralized, but rather distributed and public. 
+The identity of the participants is secured or semi-secure as it often relies on
+pseudonymity for protecting the nodes' identity, while enabling full transparency 
+in regards to the rest of the network content and operation \cite{xiao2019distributed}.
+The goal of permissionless consensus is, as generally for any consensus protocol, to reach
+agreement on a single value, or a set of values. However, due to the nature of the
+protocols, the values that are agreed upon end up establishing the serialization of the
+transactions that are executed in the network, and so establishing time consciousness and 
+total order of the events.
+
+As pointed out by \cite{xiao2019distributed} and then referenced by \cite{survey-dist-consensus}, 
+similarly to the permissioned counterparts, permissionless consensus protocols 
+aim at achieving the following properties: Termination, Agreement, Validity, 
+and Integrity. Without going into a lot of details, of these properties, 
+Agreement and Integrity are the most important ones, as they are the ones that
+guarantee the correctness of the consensus protocols. Termination and Validity
+are generally related to any classic consensus problem.
+
+\subsection{The building blocks of permissionless consensus}
+
+Also described in \cite{survey-dist-consensus}, very concisely, the 
+way to achieve an operating protocol, as seen in the mainstream blockchain
+networks, is by interchaining the generation of the agreeable value, 
+in this particular case, a block and its proof, with the dissemination of 
+the information to the rest of the network, followed by the eventual validation and 
+acceptance of the block by the rest of the nodes. This is the moment when consensus is reached.
+Nevertheless, during the whole process, a fair and somewhat predictable incentive mechanism 
+is also needed, that rewards the participants for their honest effort in reaching consensus, and
+punishes the ones that are not behaving correctly. These incentives are of major importance
+in this very context of permissionless consensus. 
+These building blocks form the basis of the inner functioning of Bitcoin itself \cite{nakamoto2008bitcoin},
+and are replicated with some variations in the other permissionless networks.
+For example, there are some networks that do not have a proof-of-work mechanism,
+but rather a proof-of-stake, or a proof-of-space, or a proof-of-burn mechanism, etc.
+when it comes to the generation of the block and its existential and verifiable proof.
+All of the other pillars are generally the same \cite{survey-dist-consensus}.
+The next sections will try to give a more detailed overview of these 
+block generation proof units.
 

### 2022-10-11 theoretical-consensus/article/article.tex
page 02

@@
 
-%\usepackage{algorithmic}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 
@@
 \hyphenation{op-tical net-works semi-conduc-tor}
+
+\algdef{SE}[SUBALG]{Indent}{EndIndent}{}{\algorithmicend\ }%
+\algtext*{Indent}
+\algtext*{EndIndent}
+
+\newcommand{\LineComment}[1]{\State \texttt{/* #1 */}}
 
@@
 Generally, consensus means reaching an agreement between multiple parties 
-in the potential presence of faulty individuals. As per multi-agent systems, 
+in the potential presence of faulty individuals. As per multiagent systems, 
 interacting over computer networks, consensus is thought to be the result 
@@
 the first solutions, not only for dealing with treason, but also for unreliable communication 
-channels, or any other kind of arbitrary byzantine behaviour, were 
+channels, or any other kind of arbitrary Byzantine behaviour, were 
 also proposed by Lamport in \cite{lamport2019byzantine,lamport1983weak}. The solution was a synchronous 
@@
 state machine" where the independent participants reach agreement not only about transactional values, 
-but also about their order - naturality forming the underlying structure of what is now known as a blockchain. 
+but also about their order - naturally forming the underlying structure of what is now known as a blockchain. 
 "Proof-of-work is essentially one-CPU-one-vote" and this is the novelty 
@@
 for permissionless consensus when there are already well known and established consensus protocols
-that work in trusted environments \cite{castro1999practical}. However, even those protocols have
-their own limitations, not only in terms of trust, fault-tolerance, centrality, permissioning,
-or bottlenecking, but also in terms of scalability \cite{miller2016honey}. Trying to put some effort on decentralization, 
+that work in trusted environments \cite{castro1999practical, miller2016honey}. However, even those protocols have
+their own limitations, not only in terms of trust, fault-tolerance, centrality, permissions,
+or bottlenecks, but also in terms of scalability \cite{miller2016honey}. Trying to put some effort on decentralization, 
 Byzantine-Fault-Tolerant consensus protocols are not known for their performance when the network grows
@@
 These requirements include dealing with today's sparse networks of anonymously and dynamically
-participating devices, without interrupting consensus and while generally battling Sybil attacks \cite{survey-dist-consensus}.
+participating devices, without interrupting consensus and while battling Sybil attacks \cite{survey-dist-consensus}.
 Fundamentally, the permissionless consensus problem is the need for a consensus protocol that
-can be run in a "zero-trust" environment, where the participants are unknown and untrusted,
+can be run in a distributed and decentralized environment, where the participants are unknown and untrusted,
 and where the network is sparse and unpredictably unreliable.
@@
 Technically, permissionless environments allow for larger networks
-that depit lower connectivity between the participants. Operationally,
+that depict lower connectivity between the participants. Operationally,
 everything is expected to happen in an asynchronous or partially synchronous fashion, 
@@
 in regards to the rest of the network content and operation \cite{xiao2019distributed}.
-The goal of permissionless consensus is, as generally for any consensus protocol, to reach
+The goal of permissionless consensus, as generally for any consensus protocol, is to reach
 agreement on a single value, or a set of values. However, due to the nature of the
@@
 way to achieve an operating protocol, as seen in the mainstream blockchain
-networks, is by interchaining the generation of the agreeable value, 
+networks, is by chaining the generation of the agreeable value, 
 in this particular case, a block and its proof, with the dissemination of 
@@
 These building blocks form the basis of the inner functioning of Bitcoin itself \cite{nakamoto2008bitcoin},
-and are replicated with some variations in the other permissionless networks.
+and are replicated with some variations in the other permissionless networks \cite{buterin2014next}.
 For example, there are some networks that do not have a proof-of-work mechanism,
 but rather a proof-of-stake, or a proof-of-space, or a proof-of-burn mechanism, etc.
-when it comes to the generation of the block and its existential and verifiable proof.
+when it comes to the generation of the block and its existential and later verifiable proof.
 All of the other pillars are generally the same \cite{survey-dist-consensus}.
@@
 block generation proof units.
+
+\section{Proof-of-Work as a reference}
+
+\subsection{The block generation}
+
+In the classical Nakamoto consensus protocol, the generation of a block, to be proposed
+for further network agreement, complies with the unit of computational work needed to
+create, or rather find, a verifiable proof of the effort spent on creating the block \cite{nakamoto2008bitcoin}. 
+This essentially requires brute forcing the finding of a cryptographic hash value for the
+aggregation of the block information with a nonce such that this hash value satisfies
+a difficulty threshold, which gets adjusted dynamically over time, to maintain the network overall 
+requirement for the block generation interval \cite{survey-dist-consensus}.
+
+Mitigation of Sybil attacks
+
+\begin{algorithm}
+\caption[short]{BlockGeneration}
+\begin{algorithmic}[1]
+  \Function {}{}
+    \State $Block Header \gets Transaction \ Merkle \ Tree \ Root$
+    \Indent
+      \State $| \ Hash \ of \ the \ last \ Block$
+      \State $| \ Timestamp$
+      \State $| \ Other$
+    \EndIndent
+
+    \LineComment{the preceding zero bits in $target$ depict the mining difficulty}
+    \While{$Hash(BlockHeader \ | \ nonce) \geq target$}
+    \State Increment $nonce$
+    \EndWhile
+    \State \Return new block
+  \EndFunction
+\end{algorithmic}
+\end{algorithm}
 

### 2022-10-11 theoretical-consensus/article/article.tex
add page 03

@@
-%% bare_jrnl.tex
+%% article.tex
 %% V1.4b
@@
 
-%\usepackage[pdftex]{graphicx}
+\usepackage[pdftex]{graphicx}
+\graphicspath{ {./images/} }
 
@@
 create, or rather find, a verifiable proof of the effort spent on creating the block \cite{nakamoto2008bitcoin}. 
-This essentially requires brute forcing the finding of a cryptographic hash value for the
-aggregation of the block information with a nonce such that this hash value satisfies
-a difficulty threshold, which gets adjusted dynamically over time, to maintain the network overall 
+This essentially requires brute forcing the search for a cryptographic hash value for the
+aggregation of the block information with a nonce, such that this hash value satisfies
+a difficulty threshold (see Algorithm \ref{alg:BlockGeneration}), which gets adjusted dynamically over time, to maintain the network overall 
 requirement for the block generation interval \cite{survey-dist-consensus}.
 
-Mitigation of Sybil attacks
+There are a couple of particularities that need to be mentioned. First, the block generation
+is of probablistic nature, as the nonce is a random value, and the hash function is a one-way function.
+Allied to this, more hashing power is not easily obtained, which allows for tackling Sybil attacks,
+despite the permissionless and pseudonymity nature of the network. Plus, the incentive mechanism
+also plays a role in this regard, by naturally incentivizing honest participation.
+The block generation interval, forged by the adjustment of the difficulty
+threshold, is also of major importance because it reduces the fork probability and
+allows for the timely propagation of proposed blocks, guaranteeing consensus
+in the presence of an honest majority of the hashing power \cite{garay2015bitcoin}.
+This probabilitic finality shifts the 1/3 BFT threshold to a 1/2 threshold.
 
 \begin{algorithm}
-\caption[short]{BlockGeneration}
+\caption[short]{BlockGeneration}\label{alg:BlockGeneration}
 \begin{algorithmic}[1]
@@
 \end{algorithm}
+
+\subsection{Trade-offs and trillema}
+
+Despite the harmony of all the aforementioned clever characteristics,
+the proof-of-work mechanism has some well known drawbacks. 
+First, the block generation is computationally expensive, and so the
+energy consumption associated with it is also high. This is caused by the
+dynamically adjusted difficulty threshold, which is a function of the block 
+generation interval. To maintain the 10 minutes interval, the hashing difficulty
+increases, and so too the hashing power and the energy needed for it. 
+However, reducing the block generation interval would increase the
+probability of forks, which results in a more frequent waste of useful work,
+loosening the security of the network. Increasing the block size would also
+have the same security effect due to the increased propagation delay.
+
+As pointed in the literature, the design of these consensus mechanisms
+shall aim for a protocular choice between a set of properties that form a trillema (see Fig. \ref{fig:trillema}): 
+Security, Scalability, and Decentralization \cite{survey-dist-consensus}.
+Summarizing, ...
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\columnwidth]{trillema}
+  \caption{From \cite{survey-dist-consensus}, the blockchain consensus trillema.}
+  \label{fig:trillema}
+\end{figure}
 

### 2022-10-12 theoretical-consensus/article/article.tex
add page 03

@@
 Generally, consensus means reaching an agreement between multiple parties 
-in the potential presence of faulty individuals. As per multiagent systems, 
+in the potential presence of faulty individuals. As per multi-agent systems, 
 interacting over computer networks, consensus is thought to be the result 
@@
 delay is bounded. However, these initial attempts failed to cover the different types of 
-faulty behaviour. Along with the establishment of the famous Byzantine Generals Problem, 
+faulty behavior. Along with the establishment of the famous Byzantine Generals Problem, 
 the first solutions, not only for dealing with treason, but also for unreliable communication 
-channels, or any other kind of arbitrary Byzantine behaviour, were 
+channels, or any other kind of arbitrary Byzantine behavior, were 
 also proposed by Lamport in \cite{lamport2019byzantine,lamport1983weak}. The solution was a synchronous 
@@
 pseudonymity for protecting the nodes' identity, while enabling full transparency 
-in regards to the rest of the network content and operation \cite{xiao2019distributed}.
+in regard to the rest of the network's content and operation \cite{xiao2019distributed}.
 The goal of permissionless consensus, as generally for any consensus protocol, is to reach
@@
 when it comes to the generation of the block and its existential and later verifiable proof.
-All of the other pillars are generally the same \cite{survey-dist-consensus}.
+All the other pillars are generally the same \cite{survey-dist-consensus}.
 The next sections will try to give a more detailed overview of these 
@@
 \subsection{The block generation}
-
-In the classical Nakamoto consensus protocol, the generation of a block, to be proposed
-for further network agreement, complies with the unit of computational work needed to
-create, or rather find, a verifiable proof of the effort spent on creating the block \cite{nakamoto2008bitcoin}. 
-This essentially requires brute forcing the search for a cryptographic hash value for the
-aggregation of the block information with a nonce, such that this hash value satisfies
-a difficulty threshold (see Algorithm \ref{alg:BlockGeneration}), which gets adjusted dynamically over time, to maintain the network overall 
-requirement for the block generation interval \cite{survey-dist-consensus}.
-
-There are a couple of particularities that need to be mentioned. First, the block generation
-is of probablistic nature, as the nonce is a random value, and the hash function is a one-way function.
-Allied to this, more hashing power is not easily obtained, which allows for tackling Sybil attacks,
-despite the permissionless and pseudonymity nature of the network. Plus, the incentive mechanism
-also plays a role in this regard, by naturally incentivizing honest participation.
-The block generation interval, forged by the adjustment of the difficulty
-threshold, is also of major importance because it reduces the fork probability and
-allows for the timely propagation of proposed blocks, guaranteeing consensus
-in the presence of an honest majority of the hashing power \cite{garay2015bitcoin}.
-This probabilitic finality shifts the 1/3 BFT threshold to a 1/2 threshold.
 
@@
 
-\subsection{Trade-offs and trillema}
+In the classical Nakamoto consensus protocol, the generation of a block, to be proposed
+for further network agreement, complies with the unit of computational work needed to
+create, or rather find, a verifiable proof of the effort spent on assembling the block \cite{nakamoto2008bitcoin}. 
+This essentially requires brute forcing the search for a cryptographic hash value for the
+aggregation of the block information with a nonce, such that this hash value satisfies
+a difficulty threshold (see Algorithm \ref{alg:BlockGeneration}), which gets adjusted dynamically over time, to maintain the network overall 
+requirement for the block generation interval \cite{survey-dist-consensus}.
+
+There are a couple of particularities that need to be mentioned. First, the block generation
+is of probabilistic nature, as the nonce is a random value, and the hash function is a one-way function.
+Allied to this, more hashing power is not easily obtained, which allows for tackling Sybil attacks,
+despite the permissionless and pseudonymity nature of the network. Plus, the incentive mechanism
+also plays a role in this regard, by naturally incentivizing honest participation.
+The block generation interval, forged by the adjustment of the difficulty
+threshold, is also of major importance because it reduces the fork probability and
+allows for the timely propagation of proposed blocks, guaranteeing consensus
+in the presence of an honest majority of the hashing power \cite{garay2015bitcoin}.
+This probabilistic finality shifts the 1/3 BFT threshold to a 1/2 threshold.
+
+\subsection{Trade-offs and trilemma}
 
@@
 dynamically adjusted difficulty threshold, which is a function of the block 
-generation interval. To maintain the 10 minutes interval, the hashing difficulty
-increases, and so too the hashing power and the energy needed for it. 
+generation interval. For example, in the Bitcoin network, to maintain the 10 
+minutes interval, the hashing difficulty
+increases with the increase of the hashing power, and so too the energy needed for it. 
 However, reducing the block generation interval would increase the
@@
 As pointed in the literature, the design of these consensus mechanisms
-shall aim for a protocular choice between a set of properties that form a trillema (see Fig. \ref{fig:trillema}): 
-Security, Scalability, and Decentralization \cite{survey-dist-consensus}.
-Summarizing, ...
+shall aim for a protocolar choice between a set of properties that form a trilemma (see Fig. \ref{fig:trilemma}): 
+Security, Scalability, and Decentralization.
+Summarizing, relaxing the security requirements may allow for more scalability, both of which,
+consequently, have hands tied with decentralization. These trade-offs are of practical
+consideration when defining the network goals and use cases \cite{survey-dist-consensus}.
 
@@
   \centering
-  \includegraphics[width=\columnwidth]{trillema}
-  \caption{From \cite{survey-dist-consensus}, the blockchain consensus trillema.}
-  \label{fig:trillema}
+  \includegraphics[width=\columnwidth]{trilemma}
+  \caption{From \cite{survey-dist-consensus}, the blockchain consensus trilemma and its inner relations.}
+  \label{fig:trilemma}
 \end{figure}

### 2022-10-22 theoretical-consensus/article/article.tex
add page 04

@@
 \begin{abstract}
-...
+Disclaimer: This paper is a work in progress. It is not a final version and it is not intended to be submited yet to the Distributed Systems Seminar. The content of this paper is subject to change.
 \end{abstract}
@@
 interacting over computer networks, consensus is thought to be the result 
-of a coordination effort such that those parties agree on some value at a 
+of a coordination effort, such that those parties agree on some value at a 
 given moment. Achieving consensus implies that the system shall be reliable 
@@
 
-The first asynchronous consensus algorithm was later proposed by Castro and Liskov in \cite{castro1999practical}. 
+The first practical asynchronous consensus algorithm was later proposed by Castro and Liskov in \cite{castro1999practical}. 
 And naturally, after that work, many other asynchronous consensus algorithms have appeared. 
@@
 can be run in a distributed and decentralized environment, where the participants are unknown and untrusted,
-and where the network is sparse and unpredictably unreliable.
+and where the network is bigger, sparse and unpredictably unreliable.
 
@@
 way to achieve an operating protocol, as seen in the mainstream blockchain
-networks, is by chaining the generation of the agreeable value, 
-in this particular case, a block and its proof, with the dissemination of 
-the information to the rest of the network, followed by the eventual validation and 
+networks, is by first generating the agreeable value, 
+in this particular case, a block and its proof, disseminating the information 
+to the rest of the network, followed by the eventual validation and 
 acceptance of the block by the rest of the nodes. This is the moment when consensus is reached.
@@
 aggregation of the block information with a nonce, such that this hash value satisfies
-a difficulty threshold (see Algorithm \ref{alg:BlockGeneration}), which gets adjusted dynamically over time, to maintain the network overall 
+a difficulty threshold (See Algorithm \ref{alg:BlockGeneration}), which gets adjusted dynamically over time, to maintain the network overall 
 requirement for the block generation interval \cite{survey-dist-consensus}.
@@
 is of probabilistic nature, as the nonce is a random value, and the hash function is a one-way function.
-Allied to this, more hashing power is not easily obtained, which allows for tackling Sybil attacks,
+Allied to this, increasing the hashing power is not an easy task, which consequently allows for tackling Sybil attacks,
 despite the permissionless and pseudonymity nature of the network. Plus, the incentive mechanism
@@
 the proof-of-work mechanism has some well known drawbacks. 
-First, the block generation is computationally expensive, and so the
+First, the block generation is computationally expensive, and so, the
 energy consumption associated with it is also high. This is caused by the
@@
 \end{figure}
+
+\section{Alternatives to Proof-of-Work}
+
+\subsection{Proof-of-Stake based protocols}
+
+At this point, we can exercise the imagination and extrapolate the
+previous block generation mechanism to a \emph{Proof-of-Something} 
+pseudo-random competition in which an entity in possession of a higher
+amount of a certain resource, either computational power, or stake, 
+or certain currency, or a higher amount of storage space, etc., guarantees
+a higher probability of leading the block generation and proposal, and
+consequently winning acceptance by the majority. This is the essence of
+the Proof-of-Stake consensus protocol, as a derivative of the Proof-of-Work
+mechanism, where a stake is a traceable and verifiable amount of a certain
+unit, token or currency, that is owned by a certain entity that wishes to
+participate in the consensus protocol. The stake works as a form of collateral
+that is used to guarantee the honesty of the entities, in an attempt to
+reduce the Sybil attack probability. And, generally as in Proof-of-Work, 
+the higher the stake, the higher the probability of leading the block generation 
+and proposal. Proof-of-Stake is, indeed, a more energy efficient consensus protocol,
+however, this adds up to nothing, especially when compared to the already proved
+BFT solutions, if centralization is a functional requirement, 
+or unfair advantage is a potential trait. 
+These limitations, along with other security concerns, may push away the 
+participation of resourceless entities, and consequently, the network
+decentralization.
+
+\subsubsection{Chain-based and Committee-based protocols}
+
+Due to the self-contained nature of this type of protocols, where the
+attestation of the stake is done inside the network, multiple classes
+of Proof-of-Stake can be distinguished, by the way the meaningfulness of
+the stake is defined and employed. For example, there are the simplest forms of \emph{PoS} that inherit most
+of the characteristics of the Proof-of-Work mechanism, but with the
+stake as part of the block generation and hashing proof (See Algorithm \ref{alg:PoSBlockGeneration}).
+These are commonly referred to as Chain-based \emph{PoS}, inheriting also the
+fault-tolerance properties of the Nakamoto consensus protocol.
+Committee-based \emph{PoS} protocols, on the other hand, are more complex,
+and they are based on the idea of a committee of entities that are 
+selected to lead the block generation and proposal. The committee is
+formed by a subset of the network entities, and the selection is done
+based on their stake. Usually, the selection takes a form of a random
+sequence of stake holders, where the probability of being selected
+for every spot is proportional to the stake. That sequence is then 
+consciously timed to orderly pace the block generation and proposal, 
+by the current slot leader. Examples of this type of mechanisms are
+the \emph{Ouroboros} \cite{kiayias2017ouroboros}, \emph{Ouroboros Praos} \cite{david2018ouroboros} protocols
+or \emph{Snow White} \cite{daian2019snow}. The longest chain rule is
+still applied, with probabilistic finality, and the 1/2 threshold is
+still required or tolerated for the network to reach consensus. Nonetheless,
+Committee-based \emph{PoS} protocols may not pair well with decentralization and 
+increase in the committee size, which detereorates the performance of the
+protocol. Several improvements have been proposed to tackle not only this
+issue, but also the security of the committees which, smaller, could be more
+vulnerable to targeted attacks \cite{david2018ouroboros}.
+
+\begin{algorithm}
+  \caption[short]{PoSBlockGeneration}\label{alg:PoSBlockGeneration}
+  \begin{algorithmic}[1]
+    \Function {}{}
+      \State $Block Header \gets Transaction \ Merkle \ Tree \ Root$
+      \Indent
+        \State $| \ Hash \ of \ the \ last \ Block$
+        \State $| \ Timestamp$
+        \State $| \ Other$
+      \EndIndent
+
+      \LineComment{the preceding zero bits in $target$ depict the mining difficulty}
+      \While{$Hash(BlockHeader \ | \ clockTime) \geq target \times stake $}
+      \State Increment $clockTime$ by a constant $tickInterval$
+      \EndWhile
+      \State \Return new block
+    \EndFunction
+  \end{algorithmic}
+\end{algorithm}
+
+\subsubsection{BFT-based protocols}
+
+These aim at achieving deterministic finality, by employing a BFT
+consensus protocol, not at the block proposal level, but at the
+block acceptance level. The idea is to have a free flow of blocks
+proposed by the network, but to have a quorum of entities that
+are responsible for the block acceptance. The block proposal mechanism
+can be of any kind, for example, Chain-based \emph{PoS}, or Committee-based \emph{PoS},
+or even pure Proof-of-Work. The block acceptance layer then allows
+for a deterministic acceptance of the blocks, by having a sort of permissioned
+BFT consensus protocol, with the trivial 1/3 byzantine tolerance. The most well known examples
+and variations are the Tendermint \cite{buchman2016tendermint} and the new Ethereum Casper 
+\cite{buterin2017casper} protocols.
+
+Delegated Proof-of-Stake (DPoS) is a variation of the BFT-based protocols,
+where the consensus group is formed by a subset of the network entities,
+which are selected by the network, with a voting mechanism based on the
+delegation of stake. The consensus is then achieved using a typical 
+PBFT protocol, with the usual 1/3 byzantine tolerance.
+
+\subsection{}
 

### 2022-10-22 theoretical-consensus/article/article.tex
add page 04

@@
 a higher probability of leading the block generation and proposal, and
-consequently winning acceptance by the majority. This is the essence of
+consequently winning the acceptance by the majority. This is the essence of
 the Proof-of-Stake consensus protocol, as a derivative of the Proof-of-Work
@@
 These limitations, along with other security concerns, may push away the 
-participation of resourceless entities, and consequently, the network
-decentralization.
+participation of resourceless entities, and consequently, affect the network
+decentralization and security.
 
@@
 Committee-based \emph{PoS} protocols, on the other hand, are more complex,
-and they are based on the idea of a committee of entities that are 
+and based on the idea of a committee of entities that are 
 selected to lead the block generation and proposal. The committee is
 formed by a subset of the network entities, and the selection is done
-based on their stake. Usually, the selection takes a form of a random
+based on their stake. Usually, the selection takes the form of a random
 sequence of stake holders, where the probability of being selected
@@
 
-\subsection{}
+\subsection{Proof-of-Something Else}
+
+Idealized and inspired by Proof-of-Stake, extending or adapting Proof-of-Work
+became a popular trend in the blockchain community. The main idea is to
+replace the computational power with some other resource, that is
+more scarce, or more valuable, or more verifiable, or more traceable, etc.,
+or to combine multiple resources, or even to add extra requirements to 
+pure Proof-of-Work. From Proof-of-Meaningful-Work, to Proof-of-Work-Time, there
+are countless and exotic variations of these mechanisms, and the list is growing \cite{token-economy-gitbook}.
+
+\subsubsection{Proof-of-Space} established itself as a popular alternative among those.
+The goal is to replace computational power with storage space,
+and if possible give it a meaningful usage. Proof-of-Retrievability is
+a cited example of this type of protocols, where the storage space is
+used to store a large amount of relevant or public data, and the retrieval of that data
+is used as a proof of the storage usage \cite{juels2007pors}. Despite the reciclability 
+of the storage space, as a resource that could be a good alternative
+to computational power, this algorithm has its own fatal flaws, depending
+mostly on a central dealer for randomness as well as for consensus fairness.
+Other reported Proof-of-Space variants may not even attempt to store/retrieve any meaningful data.
+
+\subsubsection{Proof-of-Authority} another well known derivative,
+which, in effect, is a form of Proof-of-Stake, where the stake is the 
+identity of the entity, that has been verified, scrutinized and
+approved by the network, giving it the right to participate in the
+closed consensus group. The consensus protocol, if not a BFT protocol, 
+can follow the style of a Proof-of-Work protocol, depending on the 
+requirements of the network - a choice that may shape the fault tolerance,
+decentralization, security and performance of the network \cite{survey-dist-consensus}.
+
+\subsubsection{Proof-of-Elapsed-Time} is another alternative to Proof-of-Work,
+especially targetting the intensive mining issue, that has been proposed 
+by Intel. The idea is to simulate the mining process by using a
+randomized time delay, which is proportional to the amount of work
+that would be required to mine the block. The time delay is then used to pace the
+block generation and proposal, and the block acceptance can be performed
+using a BFT protocol \cite{olson2018sawtooth}. This solution, however,
+is not a viable alternative to other, permissioned and permissionless, approaches, 
+by holding strong and centralized hardware requirements, 
+as the computations require a trusted execution environment
+(TEE) to establish the trust among the participants. Some efforts have been made
+to combine the TEE feature with Proof-of-Stake, in order to achieve a
+more decentralized and secure protocol, but the 
+same hardware requirements are still a major issue.
+
+\subsection{Other approaches}
+
+\subsubsection{The XRP Ledger Consensus Protocol} powers the XRP Ledger in a
+low-latency, high throughput, and  BFT fashion. The network itself operates somewhat
+diferently from the other blockchain references, and treats transactions as atomic units of 
+the ledger. These transactions are collected by the nodes and relayed to the
+nodes' UNL (Unique Node List), which is a list of peers that are trusted by that current
+entity. At every epoch, multiple rounds are needed to filter transactions and vote 
+on the next state of the ledger. This voting happens at the 
+UNLs' level, which are expected to be overlapped by a certain percentage,
+in order to achieve the artificial vote threshold for globally accepting the transactions.
+That threshold defines the fault tolerance of the protocol, and is set to 80\%, which 
+bounds this protocol to a 1/5 BFT allowance \cite{schwartz2014ripple, chase2018analysis}. 
+Despite that limitation, when compared to other BFT protocols, this protocol seeks to trade 
+fault tolerance for performance, by having lower connectivity and less message exchanges. 
+However, it is still expected high synchrony within the UNLs. Given these characteristics, the protocol
+is expected to be more suitable for permissioned networks, and, in terms of network size,
+is seen as not as scalable and decentralized as the permissionless ones \cite{survey-dist-consensus}.
 

### 2022-10-25 theoretical-consensus/article/article.tex
add page 05

@@
 based on their stake. Usually, the selection takes the form of a random
-sequence of stake holders, where the probability of being selected
+sequence of stakeholders, where the probability of being selected
 for every spot is proportional to the stake. That sequence is then 
@@
 Committee-based \emph{PoS} protocols may not pair well with decentralization and 
-increase in the committee size, which detereorates the performance of the
+increase in the committee size, which deteriorates the performance of the
 protocol. Several improvements have been proposed to tackle not only this
@@
 for a deterministic acceptance of the blocks, by having a sort of permissioned
-BFT consensus protocol, with the trivial 1/3 byzantine tolerance. The most well known examples
+BFT consensus protocol, with the trivial 1/3 Byzantine tolerance. The most well known examples
 and variations are the Tendermint \cite{buchman2016tendermint} and the new Ethereum Casper 
@@
 delegation of stake. The consensus is then achieved using a typical 
-PBFT protocol, with the usual 1/3 byzantine tolerance.
+PBFT protocol, with the usual 1/3 Byzantine tolerance.
 
@@
 \subsubsection{Proof-of-Elapsed-Time} is another alternative to Proof-of-Work,
-especially targetting the intensive mining issue, that has been proposed 
+especially targeting the intensive mining issue, that has been proposed 
 by Intel. The idea is to simulate the mining process by using a
@@
 low-latency, high throughput, and  BFT fashion. The network itself operates somewhat
-diferently from the other blockchain references, and treats transactions as atomic units of 
+differently from the other blockchain references, and treats transactions as atomic units of 
 the ledger. These transactions are collected by the nodes and relayed to the
@@
 is seen as not as scalable and decentralized as the permissionless ones \cite{survey-dist-consensus}.
+
+\subsubsection{DAG structures} may diverge from the previous approaches with an envisioned 
+non-linear structure for the result of the consensus attestations. The choice resides in 
+directed acyclic graphs that should expand and physically connect, further in time, the new
+data appended to the network. The data, which is typically held in the vertices of the graph, 
+can have more or less granularity, either by storing blocks of transactions, or raw transactions directly.
+This design decision may pose new potential challenges, related to the verification and finalization of
+the consensus protocol. Currently, those two approaches are then distinguished as \emph{blockDAG} and 
+\emph{txDAG}. The block-based DAG approach diverges from the traditional blockchain structure by flexibly 
+connecting every new block to multiple previous blocks. The rest of the protocol may work in the same way 
+as the traditional Proof-of-Work, however, this new feature yields new security considerations. These protocols
+are theoretically designed for very high throughput in permissionless environments, but the security trade-offs
+are still not well-known.
+In a transaction-based DAG, every vertex holds one transaction, that is connected to previous vertices. 
+All the sibling vertices shall then represent disjoint transactions to successfully deal with history and conflict
+resolutions.
 

### 2022-10-25 theoretical-consensus/article/article.tex
add page 05

@@
 \begin{abstract}
-Disclaimer: This paper is a work in progress. It is not a final version and it is not intended to be submited yet to the Distributed Systems Seminar. The content of this paper is subject to change.
+Disclaimer: This paper is a work in progress. It is not a final version and it is not intended to be submitted yet to the Distributed Systems Seminar. The content of this paper is subject to change.
 \end{abstract}
@@
 the consensus protocol. Currently, those two approaches are then distinguished as \emph{blockDAG} and 
-\emph{txDAG}. The block-based DAG approach diverges from the traditional blockchain structure by flexibly 
+\emph{txDAG} \cite{survey-dist-consensus}. 
+The block-based DAG approach diverges from the traditional blockchain structure by flexibly 
 connecting every new block to multiple previous blocks. The rest of the protocol may work in the same way 
@@
 are theoretically designed for very high throughput in permissionless environments, but the security trade-offs
-are still not well-known.
+are still not well-known \cite{sompolinsky2018phantom}.
 In a transaction-based DAG, every vertex holds one transaction, that is connected to previous vertices. 
 All the sibling vertices shall then represent disjoint transactions to successfully deal with history and conflict
-resolutions.
+resolutions. Two well-known examples of this approach are IOTA Tangle \cite{popov2018tangle} 
+and Nano \cite{lemahieu2018nano}.
+The former is of particular interest, as it is especially designed for IoT applications, low-powered devices, and
+micro-transactions. The protocol states that every new vertex needs to be connected to two previous vertices, 
+therefore approving the parent transactions, as an incentive mechanism for honest behavior \cite{survey-dist-consensus}. 
+The users are the ones to submit transactions, generating and attaching a PoW proof to it and 
+so guaranteeing probabilistic finality with 1/2 fault tolerance. This solution may indeed pair well with 
+decentralization and scalability, however, there are known security issues related to the
+easiness of growing malicious parallel branches with little computational effort, and those issues have been 
+addressed by the introduction of an undesirable centralized solution. 
+Other interesting and worth mentioning protocols are the ones that combine the DAG structure with the BFT consensus, 
+such as the Snowflake-Avalanche protocol \cite{rocket2018snowflake}.
 

### 2022-10-26 theoretical-consensus/article/article.tex
add page 05

@@
 for decentralized systems and after proof-of-work many other consensus mechanisms have been proposed, 
-based on different consensus units, like proof-of-stake, proof-of-space, proof-of-burn, etc. The chaotic 
+based on different consensus units, like proof-of-stake, proof-of-space, proof-of-authority, etc. The chaotic 
 diversity of new consensus protocols gave also room for endless reviews, overviews and comparisons \cite{survey-dist-consensus, BAMAKAN2020113385, 8400278, 8629877, 9376868, BOURAGA2021114384}. 
@@
 can be run in a distributed and decentralized environment, where the participants are unknown and untrusted,
-and where the network is bigger, sparse and unpredictably unreliable.
+and where the network is bigger, sparser and unpredictably less reliable.
 
@@
 For example, there are some networks that do not have a proof-of-work mechanism,
-but rather a proof-of-stake, or a proof-of-space, or a proof-of-burn mechanism, etc.
+but rather a proof-of-stake, or a proof-of-space, or a proof-of-authority mechanism, etc.
 when it comes to the generation of the block and its existential and later verifiable proof.
@@
 minutes interval, the hashing difficulty
-increases with the increase of the hashing power, and so too the energy needed for it. 
+increases with the increase of the hashing power, and so too the energy needed for computing that proof. 
 However, reducing the block generation interval would increase the
@@
 attestation of the stake is done inside the network, multiple classes
-of Proof-of-Stake can be distinguished, by the way the meaningfulness of
+of Proof-of-Stake can be distinguished by the way the meaningfulness of
 the stake is defined and employed. For example, there are the simplest forms of \emph{PoS} that inherit most
@@
 or to combine multiple resources, or even to add extra requirements to 
-pure Proof-of-Work. From Proof-of-Meaningful-Work, to Proof-of-Work-Time, there
+pure Proof-of-Work. From Proof-of-Meaningful-Work, or Proof-of-Luck, 
+to Proof-of-Work-Time, going through Proof-of-Humanity, or Proof-of-Personhood, there
 are countless and exotic variations of these mechanisms, and the list is growing \cite{token-economy-gitbook}.
+Not that every one of them has a big potential for entirely solving the permissionless consensus problem,
+but each one of them may tackle different use cases where consensus needs to be reached, 
+and where different resources are available to guarantee and make the agreement happen \cite{BOURAGA2021114384}.
 
@@
 
-\subsubsection{Proof-of-Authority} another well known derivative,
+\subsubsection{Proof-of-Authority} is another well known derivative,
 which, in effect, is a form of Proof-of-Stake, where the stake is the 

### 2022-10-27 theoretical-consensus/article/article.tex
add refs

@@
 These requirements include dealing with today's sparse networks of anonymously and dynamically
-participating devices, without interrupting consensus and while battling Sybil attacks \cite{survey-dist-consensus}.
+participating devices, without interrupting consensus and while battling Sybil attacks \cite{8629877, survey-dist-consensus}.
 Fundamentally, the permissionless consensus problem is the need for a consensus protocol that
@@
 transactions that are executed in the network, and so establishing time consciousness and 
-total order of the events.
+total order of the events \cite{8629877}.
 
-As pointed out by \cite{xiao2019distributed} and then referenced by \cite{survey-dist-consensus}, 
+As pointed out by \cite{xiao2019distributed, 8629877} and then referenced by \cite{survey-dist-consensus}, 
 similarly to the permissioned counterparts, permissionless consensus protocols 
@@
 a difficulty threshold (See Algorithm \ref{alg:BlockGeneration}), which gets adjusted dynamically over time, to maintain the network overall 
-requirement for the block generation interval \cite{survey-dist-consensus}.
+requirement for the block generation interval \cite{8629877, survey-dist-consensus}.
 
@@
 loosening the security of the network. Increasing the block size would also
-have the same security effect due to the increased propagation delay.
+have the same security effect due to the increased propagation delay \cite{8629877}.
 

### 2022-10-29 theoretical-consensus/article/article.tex
add page 06

@@
 \graphicspath{ {./images/} }
+\usepackage{caption}
 
@@
 based on different consensus units, like proof-of-stake, proof-of-space, proof-of-authority, etc. The chaotic 
-diversity of new consensus protocols gave also room for endless reviews, overviews and comparisons \cite{survey-dist-consensus, BAMAKAN2020113385, 8400278, 8629877, 9376868, BOURAGA2021114384}. 
+diversity of new consensus protocols gave also room for endless reviews, overviews and comparisons \cite{survey-dist-consensus, BAMAKAN2020113385, 8400278, natoli2019deconstructing, 8629877, 9376868, BOURAGA2021114384}.
 The authors of these surveys often put multiple dimensions into comparison, like fault tolerance, scalability, or energy consumption,
@@
 allows for the timely propagation of proposed blocks, guaranteeing consensus
-in the presence of an honest majority of the hashing power \cite{garay2015bitcoin}.
+in the presence of an honest majority of the hashing power \cite{garay2015bitcoin, natoli2019deconstructing}.
 This probabilistic finality shifts the 1/3 BFT threshold to a 1/2 threshold.
@@
 loosening the security of the network. Increasing the block size would also
-have the same security effect due to the increased propagation delay \cite{8629877}.
+have the same security effect due to the increased propagation delay \cite{8629877, natoli2019deconstructing}.
 
 As pointed in the literature, the design of these consensus mechanisms
-shall aim for a protocolar choice between a set of properties that form a trilemma (see Fig. \ref{fig:trilemma}): 
+shall aim for a protocolar choice between a set of properties that form a trilemma (See Fig. \ref{fig:trilemma}): 
 Security, Scalability, and Decentralization.
@@
 to Proof-of-Work-Time, going through Proof-of-Humanity, or Proof-of-Personhood, there
-are countless and exotic variations of these mechanisms, and the list is growing \cite{token-economy-gitbook}.
+are countless and exotic variations of these mechanisms, and the list is growing \cite{token-economy-gitbook, 9376868}.
 Not that every one of them has a big potential for entirely solving the permissionless consensus problem,
 but each one of them may tackle different use cases where consensus needs to be reached, 
-and where different resources are available to guarantee and make the agreement happen \cite{BOURAGA2021114384}.
+and where different resources are available to make the agreement happen \cite{BOURAGA2021114384, 9376868}.
 
@@
 
-\subsubsection{Proof-of-Elapsed-Time} is another alternative to Proof-of-Work,
+\subsubsection{Proof-of-Elapsed-Time} is another alternative to \emph{PoW},
 especially targeting the intensive mining issue, that has been proposed 
@@
 same hardware requirements are still a major issue.
+
+\subsubsection{Proof-of-Location} may be an alternative for envisioning a future where the
+consensus is achieved by taking advantage of the dynamism and easiness of movement, placement and 
+tracking of smaller and more location-versatile devices, such as mobile phones, IoT devices, etc.
+The idea behind this approach is to use the location and the movement of the devices 
+as a proof for the consensus. It relies on the location validation by close proximity, 
+for example, using low-range communication channels, either by proving the location
+of the device at a certain point in time, or by proving their proximity to other devices.
+Incentive mechanisms may play an important role to prevent location spoofing, collusion, etc. \cite{natoli2019deconstructing}.
+This protocol is getting increased attention, especially in the context of IoT, 
+with the provision of decentralized, permissionless and autonomous networks of 
+low-powered devices that can be used for providing these location verification services \cite{9376868}.
+
+\begin{figure*}[h]
+  \centering
+  \captionsetup{justification=centering}
+  \includegraphics[width=\textwidth]{proof-alternative-consensus}
+  \caption{From \cite{9376868}, an extensive list of pure proof alternative consensus mechanisms.}
+  \label{fig:proof-alternative-consensus}
+\end{figure*}
 
@@
 therefore approving the parent transactions, as an incentive mechanism for honest behavior \cite{survey-dist-consensus}. 
-The users are the ones to submit transactions, generating and attaching a PoW proof to it and 
+The users are the ones to submit transactions, generating and attaching a \emph{PoW} proof to it and 
 so guaranteeing probabilistic finality with 1/2 fault tolerance. This solution may indeed pair well with 

### 2022-10-31 theoretical-consensus/article/article.tex
add presentation

@@
 tracking of smaller and more location-versatile devices, such as mobile phones, IoT devices, etc.
-The idea behind this approach is to use the location and the movement of the devices 
+The idea behind this approach is to use the location and movement of these devices 
 as a proof for the consensus. It relies on the location validation by close proximity, 
@@
 Incentive mechanisms may play an important role to prevent location spoofing, collusion, etc. \cite{natoli2019deconstructing}.
-This protocol is getting increased attention, especially in the context of IoT, 
+These protocols are getting increased attention, especially in the context of IoT, 
 with the provision of decentralized, permissionless and autonomous networks of 
-low-powered devices that can be used for providing these location verification services \cite{9376868}.
+low-powered devices that can be used for providing location verification services \cite{9376868}.
 
@@
 fault tolerance for performance, by having lower connectivity and less message exchanges. 
-However, it is still expected high synchrony within the UNLs. Given these characteristics, the protocol
-is expected to be more suitable for permissioned networks, and, in terms of network size,
+However, it is still expected high synchrony within the UNLs. Given these characteristics, this protocol
+may be more suitable for permissioned networks, and, in terms of network size,
 is seen as not as scalable and decentralized as the permissionless ones \cite{survey-dist-consensus}.

### 2022-11-06 theoretical-consensus/article/article.tex
add page 07

@@
 The authors of these surveys often put multiple dimensions into comparison, like fault tolerance, scalability, or energy consumption,
-and, among those, some focused their efforts on mechanisms that may work in resource-constrained networks \cite{9104713, 8661654, 8168250, SALIMITARI2020100212}. 
+and, among those, some focused their efforts on mechanisms that may work in resource-constrained networks \cite{9104713, 8168250, queralta2021blockchain, SALIMITARI2020100212}. 
 This paper will try to identify the common conclusions from these comparisons, while looking into the state-of-the-art and novel approaches for running permissionless
@@
 
+\section{Resource-Constrained Networks}
+
+\subsection{Characteristics}
+
+Resource-constrained networks are a class of networks that are characterized by the
+limited access to resources, such as energy, bandwidth, storage, etc.
+These networks are typically seen in IoT applications, where the devices are
+usually low-powered, with limited computational resources, but are expected to 
+collectively operate with high availability, at least the network as a whole, 
+preferably without the need for frequent maintenance.
+These networks are also expected to be dynamic, with frequent changes in the network
+topology, by the unpredictable addition and removal of nodes. 
+The nodes are also expected to be heterogeneous, with different capabilities, and may be
+connected to the network through different communication channels. 
+Mobility may also be a feature, as the nodes may spatially move around, 
+and may directly pair with different nodes at different moments.
+The nodes may also be individually unreliable, untrusted, uncoordinated, unmanageable, may fail at any time, 
+and may be disconnected from the rest of the network.
+
+Nowadays, the typical resource-constrained networks follow a centralized architecture
+with a client-server model, where the network is controlled by a single entity,
+which is responsible for the management of the nodes, including authentication,
+authorization, and communication. This single-point-of-failure is also a major
+scalability bottleneck, and a maintenance burden, as the network grows in number of devices \cite{SALIMITARI2020100212}.
+
+\subsection{Challenges}
+
+Having in mind these characteristics of resource-constrained networks,
+and the drawbacks of the centralized architecture, it may be postulated the 
+need for a decentralized or permissionless shift that addresses some the most important challenges
+of these deployments \cite{SALIMITARI2020100212}:
+
+\begin{itemize}
+  \item \emph{Scalability}:  in terms of the number of available devices - decentralizing may allow
+  for better, faster and less costly maintenance of the network, assuring more availability and reliability
+  and cutting the single-entity dependency;
+  \item \emph{Security}: in terms of the network as a whole - removing the single-point-of-failure 
+  may reduce the attack surface, and may also allow or require more flexible and dynamic security policies;
+  \item \emph{Privacy}: in terms of the network data exchange features - decentralizing may allow for
+  a user-centric approach, where the users are in control of the data that is shared with the network; 
+\end{itemize}
+
+Permissionless consensus is a promising solution for these challenges, as an enabler for
+decentralized environments. The use cases for these networks are diverse. Nevertheless,
+as long as consensus is achieved, the network as a whole can be abstractly seen and used
+as a trusted entity that runs as a trustless decentralized service. Applications may then
+be built on top of it, and the network may be used as a platform for other services \cite{queralta2021blockchain}.
+However, there are some key aspects that need to be addressed
+in order to make this approach viable for resource-constrained networks, as we saw in the previous section:
+
+\begin{itemize}
+  \item \emph{Resource consumption}: the consensus protocols should be resource-efficient, for example, in terms of energy, bandwidth, or storage, 
+  as the nodes are expected to be low-powered devices, with limited computational, communication, and storage resources;
+  \item \emph{Heterogeneity}: the consensus protocols should account for different device characteristics, 
+  as the nodes are expected to be heterogeneous in terms of their capabilities;
+  \item \emph{Fault tolerance}: the consensus protocols should be fault-tolerant, as the nodes are expected to be unreliable, 
+  untrusted, uncoordinated, or unmanageable;
+  \item \emph{Openness}: the consensus protocols should be open and trustless, as the nodes are expected to be 
+  dynamically added and removed from the network;
+  \item \emph{Security}: the consensus protocols should be resistant to a multitude of new attacks,
+   especially if it turns out to be easy for malicious entities to overcome resource constraints;
+\end{itemize}
+
+The overall theoretical properties of the consensus protocols still need to be taken into account,
+as the goal is to have a proved functional mechanism that achieves Termination, Agreement, Validity, and Integrity,
+while playing with the trade-offs imposed by the trilemma (See Fig. \ref{fig:trilemma}) \cite{survey-dist-consensus}.
+
+\subsection{Related Work}
+
+There have been several attempts to address the challenges of resource-constrained networks,
+and to understand the limitations of the existing consensus protocols in this context.
+A common conclusion is that the existing protocols are not suitable for these networks,
+due either to the high resource consumption, or to the centralization of various aspects of the protocol,
+or also to high communication overheads, that are translated to high latency and low throughput \cite{SALIMITARI2020100212},
+making these protocols unsuitable for resource-constrained applications. Other concerns are the
+security issues, as some attempts to overcome the resource constraints may lead to new vulnerabilities,
+which may be exploited by malicious and more powerful entities.
+
+A thoroughly mentioned approach is the IOTA Tangle protocol \cite{popov2018tangle}, which
+presents a friendly design for resource-constrained networks, infinitely scalable, and with low latency.
+However, there are still known security and centralization limitations, as presented in the sections before.
+Another concept that has been explored is sharding \cite{SALIMITARI2020100212, queralta2021blockchain}, which is a
+technique that allows for the partitioning of the network into smaller sub-networks,
+that are then managed by different entities, validating transactions in parallel. 
+This approach may be used to overcome the scalability bottleneck,
+but it may also introduce new issues that are still to be researched.
+Frankly, there are indeed some well established protocols that, all in all, can be adapted to networks with
+resource-constrained devices that are not necessarily short on the resources needed to run the protocols.
+For example, a lot of \emph{PoS} related protocols may be adapted to resource-constrained networks, where 
+the stake is the only considerable resource that the nodes need. There are a multitude of projects that are
+currently exploring this approach, for example, with the Ethereum transition to \emph{PoS} \cite{buterin2017casper}.
+However, this does not address the hipothesis of taking advantage of the very nature of these limited environments, 
+as simply choosing a BFT protocol that is resource-efficient
+may also put us far apart from the ultimate decentralization and scalability goal of 
+other and fairer means of consensus participation.
+An interesting idea is based on the use of \emph{Proof of Location} for 
+coordinating the consensus. The easiness of movement and targetted location deployment
+of a lot of these low-power devices may be used to achieve a decentralized consensus
+protocol, where elected nodes may be somehow chosen based on their location, 
+either by proximity, or by some other withnessing mechanism. Variations and 
+extensions of this idea, with different and wide use cases, have been seen in
+\cite{foam2018location, xyo2018location}
+
+\section{Conclusion}
+
+Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec auctor, nisl eget aliquam aliquam.
+
+\section{Acknowledgements}
+
+This work was supported by the 2022 XRP Ledger Trust Scholarship Program for Master's Students
+at the University of Tartu.
+
 % Subsection text here.

### 2022-11-09 theoretical-consensus/article/article.tex
clean up tex

@@
 
-% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
-% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
-% *** with production work. The IEEE's font choices and paper sizes can   ***
-% *** trigger bugs that do not appear when using other class files.       ***                          ***
-% The testflow support page is at:
-% http://www.michaelshell.org/tex/testflow/
-
 \documentclass[journal]{IEEEtran}
@@
 \begin{document}
-%
-% paper title
-% Titles are generally capitalized except for words such as a, an, and, as,
-% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
-% not capitalized unless they are the first or last word of the title.
-% Line breaks \\ can be used within to get better formatting as desired.
-% Do not put math or special symbols in the title.
+
 \title{The Units of Permissionless Consensus: \\ Towards Mobile and Edge Computing}
-%
-%
-% author names and IEEE memberships
-% note positions of commas and non breaking spaces ( ~ ) LaTeX will not break
-% a structure at a ~ so this keeps an author's name from being broken across
-% two lines.
-% use \thanks{} to gain access to the first footnote area
-% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
-% was not built to handle multiple paragraphs
-%
 
@@
 
-% note the % following the last \IEEEmembership and also \thanks - 
-% these prevent an unwanted space from occurring between the last author name
-% and the end of the author line. i.e., if you had this:
-% 
-% \author{....lastname \thanks{...} \thanks{...} }
-%                     ^------------^------------^----Do not want these spaces!
-%
-% a space would be appended to the last name and could cause every name on that
-% line to be shifted left slightly. This is one of those "LaTeX things". For
-% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
-% "AB" then you have to do: "\textbf{A}\textbf{B}"
-% \thanks is no different in this regard, so shield the last } of each \thanks
-% that ends a line with a % and do not let a space in before the next \thanks.
-% Spaces after \IEEEmembership other than the last one are OK (and needed) as
-% you are supposed to have spaces between the names. For what it is worth,
-% this is a minor point as most people would not even notice if the said evil
-% space somehow managed to creep in.
-
-% The paper headers
-% \markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
-% {Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for IEEE Journals}
-% The only time the second header will appear is for the odd numbered pages
-% after the title page when using the twoside option.
-% 
-% *** Note that you probably will NOT want to include the author's ***
-% *** name in the headers of peer review papers.                   ***
-% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
-% you desire.
-
-% If you want to put a publisher's ID mark on the page you can do it like
-% this:
-%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
-% Remember, if you use this you must call \IEEEpubidadjcol in the second
-% column for its text to clear the IEEEpubid mark.
-
-% use for special paper notices
-%\IEEEspecialpapernotice{(Invited Paper)}
-
-% make the title area
 \maketitle
 
-% As a general rule, do not put math, special symbols or citations
-% in the abstract or keywords.
 \begin{abstract}
@@
 
-% Note that keywords are not normally used for peer review papers.
 \begin{IEEEkeywords}
@@
 
-% For peer review papers, you can put extra information on the cover
-% page as needed:
-% \ifCLASSOPTIONpeerreview
-% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
-% \fi
-%
-% For peerreview papers, this IEEEtran command inserts a page break and
-% creates the second title. It will be ignored for other modes.
-% \IEEEpeerreviewmaketitle
+\section{Introduction}
 
-\section{Introduction}
-% The very first letter is a 2 line initial drop letter followed
-% by the rest of the first word in caps.
-% 
-% form to use if the first word consists of a single letter:
-% \IEEEPARstart{A}{demo} file is ....
-% 
-% form to use if you need the single drop letter followed by
-% normal text (unknown if ever used by the IEEE):
-% \IEEEPARstart{A}{}demo file is ....
-% 
-% Some journals put the first two words in caps:
-% \IEEEPARstart{T}{his demo} file is ....
-% 
-% Here we have the typical use of a "T" for an initial drop letter
-% and "HIS" in caps to complete the first word.
 \IEEEPARstart{L}{ong} has been the time when consensus started 
@@
 or also to high communication overheads, that are translated to high latency and low throughput \cite{SALIMITARI2020100212},
-making these protocols unsuitable for resource-constrained applications. Other concerns are the
-security issues, as some attempts to overcome the resource constraints may lead to new vulnerabilities,
+making these protocols unsuitable for resource-constrained applications. Other concerns are
+the high space consumption, to store the older blockchain information, or the security issues, 
+as some attempts to overcome the resource constraints may lead to new vulnerabilities,
 which may be exploited by malicious and more powerful entities.
@@
 presents a friendly design for resource-constrained networks, infinitely scalable, and with low latency.
-However, there are still known security and centralization limitations, as presented in the sections before.
+However, there are still known security and centralization limitations, as presented in the sections before \cite{8168250}.
 Another concept that has been explored is sharding \cite{SALIMITARI2020100212, queralta2021blockchain}, which is a
@@
 currently exploring this approach, for example, with the Ethereum transition to \emph{PoS} \cite{buterin2017casper}.
-However, this does not address the hipothesis of taking advantage of the very nature of these limited environments, 
+However, this does not address the hypothesis of taking advantage of the very nature of these limited environments, 
 as simply choosing a BFT protocol that is resource-efficient
@@
 protocol, where elected nodes may be somehow chosen based on their location, 
-either by proximity, or by some other withnessing mechanism. Variations and 
+either by proximity, or by some other witnessing mechanism. Variations and 
 extensions of this idea, with different and wide use cases, have been seen in
@@
 
-% Subsection text here.
-
-% needed in second column of first page if using \IEEEpubid
-%\IEEEpubidadjcol
-
-% An example of a floating figure using the graphicx package.
-% Note that \label must occur AFTER (or within) \caption.
-% For figures, \caption should occur after the \includegraphics.
-% Note that IEEEtran v1.7 and later has special internal code that
-% is designed to preserve the operation of \label within \caption
-% even when the captionsoff option is in effect. However, because
-% of issues like this, it may be the safest practice to put all your
-% \label just after \caption rather than within \caption{}.
-%
-% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
-% option should be used if it is desired that the figures are to be
-% displayed while in draft mode.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=2.5in]{myfigure}
-% where an .eps filename suffix will be assumed under latex, 
-% and a .pdf suffix will be assumed for pdflatex; or what has been declared
-% via \DeclareGraphicsExtensions.
-%\caption{Simulation results for the network.}
-%\label{fig_sim}
-%\end{figure}
-
-% Note that the IEEE typically puts floats only at the top, even when this
-% results in a large percentage of a column being occupied by floats.
-
-% An example of a double column floating figure using two subfigures.
-% (The subfig.sty package must be loaded for this to work.)
-% The subfigure \label commands are set within each subfloat command,
-% and the \label for the overall figure must come after \caption.
-% \hfil is used as a separator to get equal spacing.
-% Watch out that the combined width of all the subfigures on a 
-% line do not exceed the text width or a line break will occur.
-%
-%\begin{figure*}[!t]
-%\centering
-%\subfloat[Case I]{\includegraphics[width=2.5in]{box}%
-%\label{fig_first_case}}
-%\hfil
-%\subfloat[Case II]{\includegraphics[width=2.5in]{box}%
-%\label{fig_second_case}}
-%\caption{Simulation results for the network.}
-%\label{fig_sim}
-%\end{figure*}
-%
-% Note that often IEEE papers with subfigures do not employ subfigure
-% captions (using the optional argument to \subfloat[]), but instead will
-% reference/describe all of them (a), (b), etc., within the main caption.
-% Be aware that for subfig.sty to generate the (a), (b), etc., subfigure
-% labels, the optional argument to \subfloat must be present. If a
-% subcaption is not desired, just leave its contents blank,
-% e.g., \subfloat[].
-
-% An example of a floating table. Note that, for IEEE style tables, the
-% \caption command should come BEFORE the table and, given that table
-% captions serve much like titles, are usually capitalized except for words
-% such as a, an, and, as, at, but, by, for, in, nor, of, on, or, the, to
-% and up, which are usually not capitalized unless they are the first or
-% last word of the caption. Table text will default to \footnotesize as
-% the IEEE normally uses this smaller font for tables.
-% The \label must come after \caption as always.
-%
-%\begin{table}[!t]
-%% increase table row spacing, adjust to taste
-%\renewcommand{\arraystretch}{1.3}
-% if using array.sty, it might be a good idea to tweak the value of
-% \extrarowheight as needed to properly center the text within the cells
-%\caption{An Example of a Table}
-%\label{table_example}
-%\centering
-%% Some packages, such as MDW tools, offer better commands for making tables
-%% than the plain LaTeX2e tabular which is used here.
-%\begin{tabular}{|c||c|}
-%\hline
-%One & Two\\
-%\hline
-%Three & Four\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-% Note that the IEEE does not put floats in the very first column
-% - or typically anywhere on the first page for that matter. Also,
-% in-text middle ("here") positioning is typically not used, but it
-% is allowed and encouraged for Computer Society conferences (but
-% not Computer Society journals). Most IEEE journals/conferences use
-% top floats exclusively. 
-% Note that, LaTeX2e, unlike IEEE journals/conferences, places
-% footnotes above bottom floats. This can be corrected via the
-% \fnbelowfloat command of the stfloats package.
-
-% \section{Conclusion}
-% The conclusion goes here.
-
-% if have a single appendix:
-%\appendix[Proof of the Zonklar Equations]
-% or
-%\appendix  % for no appendix heading
-% do not use \section anymore after \appendix, only \section*
-% is possibly needed
-
-% use appendices with more than one appendix
-% then use \section to start each appendix
-% you must declare a \section before using any
-% \subsection or using \label (\appendices by itself
-% starts a section numbered zero.)
-%
-
-% \appendices
-% \section{Proof of the First Zonklar Equation}
-% Appendix one text goes here.
-
-% % you can choose not to have a title for an appendix
-% % if you want by leaving the argument blank
-% \section{}
-% Appendix two text goes here.
-
-% % use section* for acknowledgment
-% \section*{Acknowledgment}
-
-% The authors would like to thank...
-
-% Can use something like this to put references on a page
-% by themselves when using endfloat and the captionsoff option.
 \ifCLASSOPTIONcaptionsoff
@@
 
-% trigger a \newpage just before the given reference
-% number - used to balance the columns on the last page
-% adjust value as needed - may need to be readjusted if
-% the document is modified later
-%\IEEEtriggeratref{8}
-% The "triggered" command can be changed if desired:
-%\IEEEtriggercmd{\enlargethispage{-5in}}
-
-% references section
-
-% can use a bibliography generated by BibTeX as a .bbl file
-% BibTeX documentation can be easily obtained at:
-% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
-% The IEEEtran BibTeX style support page is at:
-% http://www.michaelshell.org/tex/ieeetran/bibtex/
-
-%
-% <OR> manually copy in the resultant .bbl file
-% set second argument of \begin to the number of references
-% (used to reserve space for the reference number labels box)
-
 \bibliography{bib}{}
@@
 
-% biography section
-% 
-% If you have an EPS/PDF photo (graphicx package needed) extra braces are
-% needed around the contents of the optional argument to biography to prevent
-% the LaTeX parser from getting confused when it sees the complicated
-% \includegraphics command within an optional argument. (You could create
-% your own custom macro containing the \includegraphics command to make things
-% simpler here.)
-%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
-% or if you just want to reserve a space for a photo:
-
-% \begin{IEEEbiography}{Michael Shell}
-% Biography text here.
-% \end{IEEEbiography}
-
-% % if you will not have a photo at all:
-% \begin{IEEEbiographynophoto}{John Doe}
-% Biography text here.
-% \end{IEEEbiographynophoto}
-
-% insert where needed to balance the two columns on the last page with
-% biographies
-%\newpage
-
-% \begin{IEEEbiographynophoto}{Jane Doe}
-% Biography text here.
-% \end{IEEEbiographynophoto}
-
-% You can push biographies down or up by placing
-% a \vfill before or after them. The appropriate
-% use of \vfill depends on what kind of text is
-% on the last page and whether or not the columns
-% are being equalized.
-
-%\vfill
-
-% Can be used to pull up biographies so that the bottom of the last one
-% is flush with the other column.
-%\enlargethispage{-5in}
-
 % that's all folks

### 2022-11-13 practical-consensus/main.go
add in-browser-pow

@@ -1,15 +1,0 @@
-package main
-
-func main() {
-	peers := []string{}
-
-	for i := 0; i < 3; i++ {
-		peer, err := startNode(peers)
-		if err != nil {
-			panic(err)
-		}
-		peers = append(peers, peer)
-	}
-
-	select {}
-}

### 2022-11-13 practical-consensus/miner.go

@@ -1,118 +1,0 @@
-package main
-
-import (
-	"context"
-	"crypto/sha256"
-	"fmt"
-	"log"
-	"math/rand"
-	"strings"
-
-	"github.com/libp2p/go-libp2p"
-	"github.com/libp2p/go-libp2p/core/host"
-	"github.com/libp2p/go-libp2p/core/peer"
-	"github.com/libp2p/go-libp2p/core/peerstore"
-
-	ma "github.com/multiformats/go-multiaddr"
-)
-
-type Node struct {
-	Host       host.Host
-	Blockchain Blockchain
-}
-
-type Blockchain []Block
-
-var currentVal int
-
-const difficulty = 6
-
-type Block struct {
-	Val   int
-	Nonce int
-	Hash  string
-}
-
-func (b Block) String() string {
-	return fmt.Sprintf("Val: %d, Nonce: %d, Hash: %s", b.Val, b.Nonce, b.Hash)
-}
-
-func newPeer(peers []string) {
-
-	var logger = log.New(log.Writer(), "miner: ", log.Flags())
-
-	// create the miner node
-	node, err := createMinerNode(peers)
-	if err != nil {
-		logger.Printf("Error creating miner node: %v", err)
-		return
-	}
-
-	logger.SetPrefix(fmt.Sprintf("miner %v: ", node.Host.ID()))
-
-	logger.Println("info: miner node started")
-
-	// start mining
-	mine(node, logger)
-}
-
-func createMinerNode(peers []string) (Node, error) {
-	// start a libp2p node that listens on a random local TCP port
-	host, err := libp2p.New(
-		libp2p.ListenAddrStrings("/ip4/127.0.0.1/tcp/0"),
-	)
-
-	if err != nil {
-		return Node{}, fmt.Errorf("error: create node: %v", err)
-	}
-
-	node := Node{Host: host, Blockchain: Blockchain{}}
-
-	if len(peers) == 0 {
-		genesis := Block{Val: 0, Nonce: 0}
-		genesis.Hash = calculateHash(genesis)
-		node.Blockchain = append(node.Blockchain, genesis)
-	} else {
-		// TODO: connect to peers
-		for _, p := range peers {
-			paddr, err := ma.NewMultiaddr(p)
-			if err != nil {
-				return Node{}, fmt.Errorf("error: create peer address: %v", err)
-			}
-			info, err := peer.AddrInfoFromP2pAddr(paddr)
-			if err != nil {
[…]

### 2022-11-13 practical-consensus/nuxt.config.js

@@ -1,0 +1,72 @@
+export default {
+  // Disable server-side rendering: https://go.nuxtjs.dev/ssr-mode
+  ssr: false,
+
+  // Target: https://go.nuxtjs.dev/config-target
+  target: 'static',
+
+  // Global page headers: https://go.nuxtjs.dev/config-head
+  head: {
+    title: 'practical-consensus',
+    htmlAttrs: {
+      lang: 'en'
+    },
+    meta: [
+      { charset: 'utf-8' },
+      { name: 'viewport', content: 'width=device-width, initial-scale=1' },
+      { hid: 'description', name: 'description', content: '' },
+      { name: 'format-detection', content: 'telephone=no' }
+    ],
+    link: [
+      { rel: 'icon', type: 'image/x-icon', href: '/favicon.ico' }
+    ]
+  },
+
+  // Global CSS: https://go.nuxtjs.dev/config-css
+  css: [
+  ],
+
+  // Plugins to run before rendering page: https://go.nuxtjs.dev/config-plugins
+  plugins: [
+  ],
+
+  // Auto import components: https://go.nuxtjs.dev/config-components
+  components: true,
+
+  // Modules for dev and build (recommended): https://go.nuxtjs.dev/config-modules
+  buildModules: [
+    // https://go.nuxtjs.dev/typescript
+    '@nuxt/typescript-build',
+    // https://go.nuxtjs.dev/tailwindcss
+    '@nuxtjs/tailwindcss',
+  ],
+
+  // Modules: https://go.nuxtjs.dev/config-modules
+  modules: [
+  ],
+
+  // Build Configuration: https://go.nuxtjs.dev/config-build
+  build: {
+    extend(config, ctx) {
+      if (ctx.isClient) {
+        config.module.rules.push(
+          {
+            test: /\.worker\.ts$/,
+            loader: "worker-loader",
+            exclude: /(node_modules)/,
+          }, 
+          {
+            test: /\.worker\.js$/,
+            loader: 'worker-loader',
+            exclude: /(node_modules)/,
+        })
+      }
+    },
+  },
+
+  env: {
+    // comma separated
+    gunRelays: process.env.GUN_RELAYS || 'http://localhost:8765/gun',
+    difficulty: process.env.DIFFICULTY || "5",
+  }
+}

### 2022-11-13 practical-consensus/peer.go

@@ -1,157 +1,0 @@
-package main
-
-import (
-	"context"
-	"encoding/json"
-	"fmt"
-	"log"
-	"math/rand"
-	"time"
-
-	"github.com/libp2p/go-libp2p"
-	"github.com/libp2p/go-libp2p/core/host"
-	"github.com/libp2p/go-libp2p/core/peer"
-	"github.com/libp2p/go-libp2p/core/peerstore"
-
-	pubsub "github.com/libp2p/go-libp2p-pubsub"
-	ma "github.com/multiformats/go-multiaddr"
-)
-
-func startNode(peers []string) (string, error) {
-	var logger = log.New(log.Writer(), "peer: ", log.Flags())
-
-	node, err := createNode(peers)
-	if err != nil {
-		logger.Printf("error: create node: %v", err)
-		return "", err
-	}
-
-	topic, sub, err := createPubSub(node, logger)
-	if err != nil {
-		logger.Printf("error: create pubsub: %v", err)
-		return "", err
-	}
-
-	logger.SetPrefix(fmt.Sprintf("peer %v: ", node.ID().ShortString()))
-
-	logger.Println("info: peer started")
-
-	go subscribe(node, sub, logger)
-
-	go publish(node, topic, logger)
-
-	return getPeerMultiAddress(node), nil
-}
-
-func createNode(peers []string) (host.Host, error) {
-	// start a libp2p node that listens on a random local TCP port
-	node, err := libp2p.New(
-		libp2p.ListenAddrStrings("/ip4/127.0.0.1/tcp/0"),
-	)
-	if err != nil {
-		return nil, err
-	}
-
-	// add the peers to the peerstore
-	for _, p := range peers {
-		paddr, err := ma.NewMultiaddr(p)
-		if err != nil {
-			return nil, fmt.Errorf("error: create peer address: %v", err)
-		}
-		info, err := peer.AddrInfoFromP2pAddr(paddr)
-		if err != nil {
-			return nil, fmt.Errorf("error: create peer info: %v", err)
-		}
-
-		node.Peerstore().AddAddr(info.ID, paddr, peerstore.PermanentAddrTTL)
-		if err := node.Connect(context.Background(), *info); err != nil {
-			return nil, fmt.Errorf("error: connect to peer: %v", err)
-		}
-	}
-
-	return node, nil
-}
-
[…]

### 2022-11-13 practical-consensus/src/Block.ts

@@ -1,0 +1,41 @@
+import sha256 from 'crypto-js/sha256';
+
+const DIFFICULTY = parseInt(process.env.difficulty?.toString() || '5');
+const TARGET = parseInt(`0x${"0".repeat(DIFFICULTY)}${"f".repeat(64-DIFFICULTY)}`,16);
+
+// Block class
+export class Block {
+
+    public hash: string;
+    public previousHash: string;
+    public nonce: number;
+    public data: string;
+
+    constructor(hash: string, previousHash: string, nonce: number, data: string) {
+        this.hash = hash;
+        this.previousHash = previousHash;
+        this.nonce = nonce;
+        this.data = data;
+    }
+
+    public prettyPrint() {
+        return `Block:
+        Hash: ${this.hash}
+        Previous Hash: ${this.previousHash}
+        Nonce: ${this.nonce}
+        Data: ${this.data}
+        `;
+    }
+
+    static isHashValid(hash: string) {
+        return parseInt(hash, 16) < TARGET;
+    }
+
+    static calculateHash(previousHash: string, nonce: number, data: string): string {
+        return sha256(previousHash + nonce + data).toString();
+    }
+
+    static genesisBlock(): Block {
+        return new Block('0000003466f7289d52851720ad259c88f192604759e34ece83d6eb74fa03b5ed', '', 9865212, 'edurbrito');
+    }
+}

### 2022-11-13 practical-consensus/src/Blockchain.ts

@@ -1,0 +1,18 @@
+import { Block } from "./Block";
+
+export class Blockchain {
+    public activeBlocks:Map<string, Block> = new Map<string,Block>();
+    public mainChain: Block[] = [];
+
+    public addBlock(block: Block): void {
+        let parentBlock = this.activeBlocks.get(block.previousHash);
+        if (parentBlock !== undefined) {
+            this.mainChain.unshift(parentBlock);
+            this.activeBlocks.delete(block.previousHash);
+        }
+        
+        if (this.mainChain.length == 0 || this.mainChain[0].hash === block.previousHash) {
+            this.activeBlocks.set(block.hash, block);
+        }
+    }
+}

### 2022-11-13 practical-consensus/src/miner.worker.ts

@@ -1,0 +1,27 @@
+import { Block } from "./Block";
+
+const ctx: Worker = self as any;
+
+function mine(currentBlock: Block, identifier: string) {
+    let nonce = 0;
+    // create a new block
+    while(true) {
+            let hash = Block.calculateHash(currentBlock.hash, nonce, identifier);
+            if(Block.isHashValid(hash)) {
+                let block = new Block(hash, currentBlock.hash, nonce, identifier);
+                currentBlock = block;
+    
+                // send the new block to the main thread
+                ctx.postMessage(block);
+                nonce = 0;
+            }
+            nonce++;
+    }
+}
+
+// on message "mine" from the main thread
+ctx.onmessage = (e) => {
+    if (e.data[0] === "mine") {
+        mine(e.data[1], e.data[2]);
+    }
+}

### 2022-11-13 practical-consensus/src/relay/build-relay.sh

@@ -1,0 +1,5 @@
+cd /tmp
+git clone https://github.com/amark/gun.git
+cd /tmp/gun
+sudo docker build -t in-browser-pow/gundb:v1 .
+# sudo docker run -d -p 8765:8765 in-browser-pow/gundb:v1

### 2022-11-13 practical-consensus/typings/worker-loader.d.ts

@@ -1,0 +1,10 @@
+declare module "*.worker.ts" {
+    // You need to change `Worker`, if you specified a different value for the `workerType` option
+    class WebpackWorker extends Worker {
+        constructor();
+    }
+
+    // Uncomment this if you set the `esModule` option to `false`
+    // export = WebpackWorker;
+    export default WebpackWorker;
+}

### 2022-11-16 theoretical-consensus/article/article.tex
add appendix

@@
 
+\section[Appendix. In-browser Proof-of-Work]{Appendix\\ {In-browser Proof-of-Work}}
+
+Along with this research, a simple practical prototype was developed
+to demonstrate the effectiveness of Proof-of-Work as a permissionless consensus protocol.
+The prototype was developed with the idea of being accessible to any device that supports
+the modern features of web browsers. Indeed, this prototype can run in a seemingly decentralized fashion,
+being delivered as a static web page, without the inherent need for a centralized hosting or backend server.
+The next sections will give more details about the implementation and design considerations, 
+as well as a short set of remarks and future work.
+
+\subsection{Implementation}
+
+This prototype was developed and shipped as a static web page.
+The base layer tooling choice resided in the \emph{Nuxt.js} framework, a meta-framework
+for building web applications based on \emph{Vue.js}. With that established,
+it became clear the need for a protocol for syncing and communicating in a 
+peer-to-peer way between the devices. The choice resided in the \emph{GUN} ecosystem,
+which ships a small but powerful set of tools for in-browser data synchronization.
+Since there was \emph{TypeScript} support from all the major dependencies of this project,
+this was the language chosen for the interactive web development. The cryptographic standards used in the 
+\emph{PoW} algorithm were imported from the \emph{crypto-js} library. And, finally, HTML and CSS, with the help
+of the \emph{Tailwind CSS} framework were used for styling and presentation.
+
+A major requirement for such a system is the capacity for parallel computations, 
+in order to be able to, simultaneously, listen for and validate proposed blocks, 
+present the current results in the screen, and mine new blocks. The traditional 
+\emph{JavaScript} computation model is a single-threaded environment that, naturally,
+does not allow for the efficient and effective handling of all these scenarios, happening
+at the same time. Most of the modern web browsers have a living solution for this problem, 
+and that is based on \emph{Web Workers}. This \emph{API} was used in order to offload
+the expensive mining computations and take that burden out of the main thread, leaving it
+with the responsibility of listening for information updates, either from other peers, with
+proposed blocks, or from the \emph{miner} threads, with new mined blocks, and decide and update
+the user interface accordingly.
+
+The remaining piece of the puzzle is the device discovery issue. The way the internet works does not
+allow for direct device pairing and easy peer discovery. The modern web generally follows a 
+client-server model with the centralization of resources. And, for this project, there was the need
+for a mechanism that would allow the peers to connect to each other to then proceed to synchronize
+the information. The \emph{GUN} ecosystem established a commonly accepted solution for this issue, that, 
+very shortly, depends on setting up \emph{relay} peers that have somewhat static addresses and that can 
+serve as reference points and pairing gateways for the nodes to discover and establish direct communications 
+with their peers. For development and demonstration purposes, multiple relay nodes were set up, running 
+as \emph{Docker} containers, with static ports and addresses, that could then be referenced in the web 
+page generation, for the other mobile and dynamic peers to connect to and make contacts with the rest of the network.
+The communication fashion is abstracted by the libraries used, but, generally, it happens either via \emph{WebSockets}
+or \emph{WebRTC} technology. Figure \ref{fig:in-browser-pow-topology} shows the topology of the whole system.
+
+\begin{figure*}[h]
+  \centering
+  \captionsetup{justification=centering}
+  % \includegraphics{in-browser-pow-topology}
+  \caption{blblb}
+  \label{fig:in-browser-pow-topology}
+\end{figure*}
+
+\subsection{Remarks}
+
+The prototype was tested, is demonstrable and works as intended. Some \emph{Nakamoto} chain-based
+features were simplified, for the sake of maintaining a didactic complexity. For example, 
+the Longest-Chain rule, in this prototype, works by having a temporary set of active blocks 
+that get appended to the main chain if there are new blocks mined on top
+of them. This setting enforces higher synchrony between the peers, as there is no mechanism for
+general chain data synchronization. This also means that a new peer joining the network does not
+need or does not ask for the rest of the chain history to start mining, but just the most recent block.
+Again, this is justified by the purpose of this prototype which was to demonstrate the achievement
+of consensus in a permissionless environment.
+
+No extensive benchmarks or analysis were made, other than live tests with multiple
+and heterogeneous devices. The basic tests involved one, two or more browser tabs
+opened in the same device, competing for mining the blocks. The distribution of blocks mined
+in a fixed and considerable interval was observed to be uniform for the participants. Another test was
+conducted with a laptop and a mobile device, for which, for obvious reasons, the laptop
+had a generally bigger advantage in the mining process.
+
+\subsection{Future Work}
+
 % that's all folks

### 2022-11-16 theoretical-consensus/article/article.tex
add page 09

@@
 \begin{abstract}
-Disclaimer: This paper is a work in progress. It is not a final version and it is not intended to be submitted yet to the Distributed Systems Seminar. The content of this paper is subject to change.
+
+This paper presents an overview of the consensus problem
+applicable to permissionless environments. Starting from 
+the classical roots of consensus in distributed systems,
+and then transitioning to the more recent developments in the
+field of permissionless consensus, this paper also reviews
+the common approaches and architectural design of the most well known protocols, 
+and discusses the trade-offs between the different approaches.
+Finally, this paper points out the challenges that
+permissionless consensus faces in the context of resource
+constrained environments, such as mobile and edge computing,
+identifies and argues about the current approaches to address these challenges,
+and shows potential research directions.
+
 \end{abstract}
@@
 \begin{IEEEkeywords}
-IEEE, IEEEtran, journal, \LaTeX, paper, template.
+Consensus, Distributed Systems, Permissionless Consensus, Mobile and Edge Computing, Blockchain, Resource Constrained Environments.
 \end{IEEEkeywords}
@@
 
+\begin{figure*}[h]
+  \centering
+  \captionsetup{justification=centering}
+  \includegraphics[width=\textwidth]{proof-alternative-consensus}
+  \caption{From \cite{9376868}, an extensive list of pure proof alternative consensus mechanisms.}
+  \label{fig:proof-alternative-consensus}
+\end{figure*}
+
 \subsubsection{Proof-of-Location} may be an alternative for envisioning a future where the
@@
 low-powered devices that can be used for providing location verification services \cite{9376868}.
-
-\begin{figure*}[h]
-  \centering
-  \captionsetup{justification=centering}
-  \includegraphics[width=\textwidth]{proof-alternative-consensus}
-  \caption{From \cite{9376868}, an extensive list of pure proof alternative consensus mechanisms.}
-  \label{fig:proof-alternative-consensus}
-\end{figure*}
 
@@
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec auctor, nisl eget aliquam aliquam.
+In this paper, we have presented a comprehensive summary of the permissionless
+consensus problem, the most widely used consensus protocols, and the challenges
+that arise when applying these protocols to resource-constrained networks.
+We have presented the challenges that these networks are facing, and the 
+limitations of the existing consensus protocols.
+We have also presented the main approaches that have been explored to address these challenges,
+and we have showed potential paths for future research.
 
@@
 This prototype was developed and shipped as a static web page.
-The base layer tooling choice resided in the \emph{Nuxt.js} framework, a meta-framework
-for building web applications based on \emph{Vue.js}. With that established,
+The base layer tooling choice resided in the \emph{Nuxt.js} framework \cite{nuxt}, a meta-framework
+for building web applications based on \emph{Vue.js} \cite{vuejs}. With that established,
 it became clear the need for a protocol for syncing and communicating in a 
-peer-to-peer way between the devices. The choice resided in the \emph{GUN} ecosystem,
+peer-to-peer way between the devices. The choice resided in the \emph{GUN} ecosystem \cite{gundb},
 which ships a small but powerful set of tools for in-browser data synchronization.
 Since there was \emph{TypeScript} support from all the major dependencies of this project,
-this was the language chosen for the interactive web development. The cryptographic standards used in the 
-\emph{PoW} algorithm were imported from the \emph{crypto-js} library. And, finally, HTML and CSS, with the help
-of the \emph{Tailwind CSS} framework were used for styling and presentation.
+this was the language chosen for the interactive web development \cite{typescript}. The cryptographic standards used in the 
+\emph{PoW} algorithm were imported from the \emph{crypto-js} library \cite{crypto-js}. And, finally, HTML and CSS, with the help
+of the \emph{Tailwind CSS} framework \cite{tailwindcss} were used for styling and presentation.
 
@@
 at the same time. Most of the modern web browsers have a living solution for this problem, 
-and that is based on \emph{Web Workers}. This \emph{API} was used in order to offload
+and that is based on \emph{Web Workers} \cite{webworkers}. This \emph{API} was used in order to offload
 the expensive mining computations and take that burden out of the main thread, leaving it
@@
 that get appended to the main chain if there are new blocks mined on top
-of them. This setting enforces higher synchrony between the peers, as there is no mechanism for
-general chain data synchronization. This also means that a new peer joining the network does not
+of them. This setting enforces higher synchrony between the peers, as there is also no mechanism for
+general chain data synchronization. Additionally, this means that a new peer joining the network does not
 need or does not ask for the rest of the chain history to start mining, but just the most recent block.
-Again, this is justified by the purpose of this prototype which was to demonstrate the achievement
+The difficulty was also set to a constant value, for the sake of simplicity.
+Again, this is justified by the purpose of this prototype, which was to demonstrate the achievement
 of consensus in a permissionless environment.
@@
 
+The prototype was developed as a proof-of-concept and a demonstration of the feasibility
+of the \emph{PoW} consensus protocol in a permissionless environment. The next steps
+for this project would be to extend the functionality and the features of the system,
+in order to make it more robust and usable. Some of the ideas for future work are, for example,
+the improvement of the \emph{PoW} algorithm, by adding a difficulty adjustment mechanism,
+to make the mining process more fair for computationally weaker devices, or to
+add a mechanism for the synchronization of the old chain history. Another idea would be to
+improve the parallelization of the computations, by more efficiently distributing the
+workload among mining threads. Finally, the user interface could be improved 
+with more informative and interactive features, that would allow the user to have a 
+deeper understanding of the system and the consensus process.
+
 % that's all folks

### 2022-11-17 theoretical-consensus/article/article.tex
add page 10

@@
 
-\author{Eduardo Ribas Brito \\ eduardo.ribas.brito@ut.ee}
+\author{Eduardo Ribas Brito 
+\\Supervisors: Ulrich Norbisrath, Eero Vainikko
+\\Distributed Systems Seminar, Fall 2022
+\\\emph{Institute of Computer Science}
+\\\emph{University of Tartu, Estonia}
+\\ eduardo.ribas.brito@ut.ee}
 
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\columnwidth]{building-blocks-consensus.png}
+  \caption{An illustration of the permissionless consensus building blocks - 
+  the block generation and proposal phases followed by the block validation, 
+  along with frequent network topology changes and the consequent serialization of the information.}
+  \label{fig:building-blocks-consensus}
+\end{figure}
+
 \subsection{The building blocks of permissionless consensus}
@@
 to the rest of the network, followed by the eventual validation and 
-acceptance of the block by the rest of the nodes. This is the moment when consensus is reached.
+acceptance of the block by the rest of the nodes. This is the moment when consensus is reached (See Fig. \ref{fig:building-blocks-consensus}).
 Nevertheless, during the whole process, a fair and somewhat predictable incentive mechanism 
@@
 
+\begin{algorithm}
+  \caption[short]{PoSBlockGeneration}\label{alg:PoSBlockGeneration}
+  \begin{algorithmic}[1]
+    \Function {}{}
+      \State $Block Header \gets Transaction \ Merkle \ Tree \ Root$
+      \Indent
+        \State $| \ Hash \ of \ the \ last \ Block$
+        \State $| \ Timestamp$
+        \State $| \ Other$
+      \EndIndent
+
+      \LineComment{the preceding zero bits in $target$ depict the mining difficulty}
+      \While{$Hash(BlockHeader \ | \ clockTime) \geq target \times stake $}
+      \State Increment $clockTime$ by a constant $tickInterval$
+      \EndWhile
+      \State \Return new block
+    \EndFunction
+  \end{algorithmic}
+\end{algorithm}
+
 \subsubsection{Chain-based and Committee-based protocols}
@@
 
-\begin{algorithm}
-  \caption[short]{PoSBlockGeneration}\label{alg:PoSBlockGeneration}
-  \begin{algorithmic}[1]
-    \Function {}{}
-      \State $Block Header \gets Transaction \ Merkle \ Tree \ Root$
-      \Indent
-        \State $| \ Hash \ of \ the \ last \ Block$
-        \State $| \ Timestamp$
-        \State $| \ Other$
-      \EndIndent
-
-      \LineComment{the preceding zero bits in $target$ depict the mining difficulty}
-      \While{$Hash(BlockHeader \ | \ clockTime) \geq target \times stake $}
-      \State Increment $clockTime$ by a constant $tickInterval$
-      \EndWhile
-      \State \Return new block
-    \EndFunction
-  \end{algorithmic}
-\end{algorithm}
-
 \subsubsection{BFT-based protocols}
@@
 proposed by the network, but to have a quorum of entities that
-are responsible for the block acceptance. The block proposal mechanism
+are responsible for the block acceptance. The block generation and proposal mechanism
 can be of any kind, for example, Chain-based \emph{PoS}, or Committee-based \emph{PoS},
@@
 PBFT protocol, with the usual 1/3 Byzantine tolerance.
+
+\begin{figure*}[h]
+  \centering
+  \captionsetup{justification=centering}
+  \includegraphics[width=\textwidth]{proof-alternative-consensus}
+  \caption{From \cite{9376868}, an extensive list of pure proof alternative consensus mechanisms.}
+  \label{fig:proof-alternative-consensus}
+\end{figure*}
 
@@
 
-\subsubsection{Proof-of-Authority} is another well known derivative,
+\subsubsection{Proof-of-Authority} is another well known derivative
 which, in effect, is a form of Proof-of-Stake, where the stake is the 
@@
 
-\begin{figure*}[h]
-  \centering
-  \captionsetup{justification=centering}
-  \includegraphics[width=\textwidth]{proof-alternative-consensus}
-  \caption{From \cite{9376868}, an extensive list of pure proof alternative consensus mechanisms.}
-  \label{fig:proof-alternative-consensus}
-\end{figure*}
-
-\subsubsection{Proof-of-Location} may be an alternative for envisioning a future where the
+\subsubsection{Proof-of-Location} may be an alternative for envisioning a future where
 consensus is achieved by taking advantage of the dynamism and easiness of movement, placement and 
@@
 The former is of particular interest, as it is especially designed for IoT applications, low-powered devices, and
-micro-transactions. The protocol states that every new vertex needs to be connected to two previous vertices, 
+microtransactions. The protocol states that every new vertex needs to be connected to two previous vertices, 
 therefore approving the parent transactions, as an incentive mechanism for honest behavior \cite{survey-dist-consensus}. 
@@
   \item \emph{Security}: in terms of the network as a whole - removing the single-point-of-failure 
-  may reduce the attack surface, and may also allow or require more flexible and dynamic security policies;
+  may reduce the attack surface, and may also allow for more flexible and dynamic security policies;
   \item \emph{Privacy}: in terms of the network data exchange features - decentralizing may allow for
@@
 Another concept that has been explored is sharding \cite{SALIMITARI2020100212, queralta2021blockchain}, which is a
-technique that allows for the partitioning of the network into smaller sub-networks,
+technique that allows for the partitioning of the network into smaller subnetworks,
 that are then managed by different entities, validating transactions in parallel. 
@@
 An interesting idea is based on the use of \emph{Proof of Location} for 
-coordinating the consensus. The easiness of movement and targetted location deployment
+coordinating the consensus. The easiness of movement and targeted location deployment
 of a lot of these low-power devices may be used to achieve a decentralized consensus
@@
 
-In this paper, we have presented a comprehensive summary of the permissionless
-consensus problem, the most widely used consensus protocols, and the challenges
-that arise when applying these protocols to resource-constrained networks.
-We have presented the challenges that these networks are facing, and the 
-limitations of the existing consensus protocols.
+In this paper, it was presented a comprehensive summary of the permissionless
+consensus problem, along with a short analysis of the most widely used consensus 
+protocols, and the challenges that arise when applying these protocols to different environments.
+In particular, this paper presented the challenges that resource-constrained networks may be facing, 
+and the limitations of the existing consensus protocols when applied to this type of networks.
 We have also presented the main approaches that have been explored to address these challenges,
-and we have showed potential paths for future research.
+and showed some potential paths for future research.
 
@@
 
+\newpage
+
 \section[Appendix. In-browser Proof-of-Work]{Appendix\\ {In-browser Proof-of-Work}}
+
+\bigbreak
 
@@
 of the \emph{Tailwind CSS} framework \cite{tailwindcss} were used for styling and presentation.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\columnwidth]{in-browser-pow-topology}
+  \caption{The topology of the system, with an underlay network of relay servers that allow the overlay network of devices to discover their peers and establish direct connections.}
+  \label{fig:in-browser-pow-topology}
+\end{figure}
 
@@
 
-\begin{figure*}[h]
-  \centering
-  \captionsetup{justification=centering}
-  % \includegraphics{in-browser-pow-topology}
-  \caption{blblb}
-  \label{fig:in-browser-pow-topology}
-\end{figure*}
-
 \subsection{Remarks}
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\columnwidth]{in-browser-pow}
+  \caption{The user interface showing three browser tabs opened and competing for mining the next blocks, 
+  while achieving consensus and presenting the same view.}
+  \label{fig:in-browser-pow}
+\end{figure}
+
 \subsection{Future Work}
@@
 for this project would be to extend the functionality and the features of the system,
-in order to make it more robust and usable. Some of the ideas for future work are, for example,
+in order to make it more robust and usable. Some ideas for future work are, for example,
 the improvement of the \emph{PoW} algorithm, by adding a difficulty adjustment mechanism,
@@
 with more informative and interactive features, that would allow the user to have a 
-deeper understanding of the system and the consensus process.
+deeper understanding of the system and the consensus process (See Fig. \ref{fig:in-browser-pow}).
+
+\subsection{Repository}
+
+All the work developed, both the practical and the theoretical components, can be found 
+at \emph{\url{https://github.com/edurbrito/dist-sys-seminar}}.
 

### 2022-11-17 theoretical-consensus/article/article.tex
add readme

@@
 the classical roots of consensus in distributed systems,
-and then transitioning to the more recent developments in the
+transitioning to the more recent developments in the
 field of permissionless consensus, this paper also reviews
@@
 permissionless consensus faces in the context of resource
-constrained environments, such as mobile and edge computing,
-identifies and argues about the current approaches to address these challenges,
+constrained environments, such as mobile and edge computing.
+It identifies and argues about the current approaches to address these challenges,
 and shows potential research directions.
@@
 guarantee the correctness of the consensus protocols. Termination and Validity
-are generally related to any classic consensus problem.
+are generally related to any classical consensus problem.
 

### 2022-11-19 theoretical-consensus/article/article.tex
change figures placement
fix genesis block & readme

@@
 the common approaches and architectural design of the most well known protocols, 
-and discusses the trade-offs between the different approaches.
+and discusses the trade-offs between the different strategies.
 Finally, this paper points out the challenges that
@@
 
+In the classical Nakamoto consensus protocol, the generation of a block, to be proposed
+for further network agreement, complies with the unit of computational work needed to
+create, or rather find, a verifiable proof of the effort spent on assembling the block \cite{nakamoto2008bitcoin}. 
+This essentially requires brute forcing the search for a cryptographic hash value for the
+aggregation of the block information with a nonce, such that this hash value satisfies
+a difficulty threshold (See Algorithm \ref{alg:BlockGeneration}), which gets adjusted dynamically over time, to maintain the network overall 
+requirement for the block generation interval \cite{8629877, survey-dist-consensus}.
+
+There are a couple of particularities that need to be mentioned. First, the block generation
+is of probabilistic nature, as the nonce is a random value, and the hash function is a one-way function.
+Allied to this, increasing the hashing power is not an easy task, which consequently allows for tackling Sybil attacks,
+despite the permissionless and pseudonymity nature of the network. Plus, the incentive mechanism
+also plays a role in this regard, by naturally incentivizing honest participation.
+The block generation interval, forged by the adjustment of the difficulty
+threshold, is also of major importance because it reduces the fork probability and
+allows for the timely propagation of proposed blocks, guaranteeing consensus
+in the presence of an honest majority of the hashing power \cite{garay2015bitcoin, natoli2019deconstructing}.
+This probabilistic finality shifts the 1/3 BFT threshold to a 1/2 threshold.
+
 \begin{algorithm}
@@
 \end{algorithm}
-
-In the classical Nakamoto consensus protocol, the generation of a block, to be proposed
-for further network agreement, complies with the unit of computational work needed to
-create, or rather find, a verifiable proof of the effort spent on assembling the block \cite{nakamoto2008bitcoin}. 
-This essentially requires brute forcing the search for a cryptographic hash value for the
-aggregation of the block information with a nonce, such that this hash value satisfies
-a difficulty threshold (See Algorithm \ref{alg:BlockGeneration}), which gets adjusted dynamically over time, to maintain the network overall 
-requirement for the block generation interval \cite{8629877, survey-dist-consensus}.
-
-There are a couple of particularities that need to be mentioned. First, the block generation
-is of probabilistic nature, as the nonce is a random value, and the hash function is a one-way function.
-Allied to this, increasing the hashing power is not an easy task, which consequently allows for tackling Sybil attacks,
-despite the permissionless and pseudonymity nature of the network. Plus, the incentive mechanism
-also plays a role in this regard, by naturally incentivizing honest participation.
-The block generation interval, forged by the adjustment of the difficulty
-threshold, is also of major importance because it reduces the fork probability and
-allows for the timely propagation of proposed blocks, guaranteeing consensus
-in the presence of an honest majority of the hashing power \cite{garay2015bitcoin, natoli2019deconstructing}.
-This probabilistic finality shifts the 1/3 BFT threshold to a 1/2 threshold.
 
@@
 
-\begin{algorithm}
-  \caption[short]{PoSBlockGeneration}\label{alg:PoSBlockGeneration}
-  \begin{algorithmic}[1]
-    \Function {}{}
-      \State $Block Header \gets Transaction \ Merkle \ Tree \ Root$
-      \Indent
-        \State $| \ Hash \ of \ the \ last \ Block$
-        \State $| \ Timestamp$
-        \State $| \ Other$
-      \EndIndent
-
-      \LineComment{the preceding zero bits in $target$ depict the mining difficulty}
-      \While{$Hash(BlockHeader \ | \ clockTime) \geq target \times stake $}
-      \State Increment $clockTime$ by a constant $tickInterval$
-      \EndWhile
-      \State \Return new block
-    \EndFunction
-  \end{algorithmic}
-\end{algorithm}
-
 \subsubsection{Chain-based and Committee-based protocols}
@@
 PBFT protocol, with the usual 1/3 Byzantine tolerance.
+
+\begin{algorithm}
+  \caption[short]{PoSBlockGeneration}\label{alg:PoSBlockGeneration}
+  \begin{algorithmic}[1]
+    \Function {}{}
+      \State $Block Header \gets Transaction \ Merkle \ Tree \ Root$
+      \Indent
+        \State $| \ Hash \ of \ the \ last \ Block$
+        \State $| \ Timestamp$
+        \State $| \ Other$
+      \EndIndent
+
+      \LineComment{the preceding zero bits in $target$ depict the mining difficulty}
+      \While{$Hash(BlockHeader \ | \ clockTime) \geq target \times stake $}
+      \State Increment $clockTime$ by a constant $tickInterval$
+      \EndWhile
+      \State \Return new block
+    \EndFunction
+  \end{algorithmic}
+\end{algorithm}
 
@@
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\columnwidth]{in-browser-pow-topology}
-  \caption{The topology of the system, with an underlay network of relay servers that allow the overlay network of devices to discover their peers and establish direct connections.}
-  \label{fig:in-browser-pow-topology}
-\end{figure}
-
 A major requirement for such a system is the capacity for parallel computations, 
@@
 The communication fashion is abstracted by the libraries used, but, generally, it happens either via \emph{WebSockets}
-or \emph{WebRTC} technology. Figure \ref{fig:in-browser-pow-topology} shows the topology of the whole system.
+or \emph{WebRTC} technology. Fig. \ref{fig:in-browser-pow-topology} shows the topology of the whole system.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\columnwidth]{in-browser-pow-topology}
+  \caption{The topology of the system, with an underlay network of relay servers that allow the overlay network of devices to discover their peers and establish direct connections.}
+  \label{fig:in-browser-pow-topology}
+\end{figure}
 
@@
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\columnwidth]{in-browser-pow}
-  \caption{The user interface showing three browser tabs opened and competing for mining the next blocks, 
-  while achieving consensus and presenting the same view.}
-  \label{fig:in-browser-pow}
-\end{figure}
-
 \subsection{Future Work}
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\columnwidth]{in-browser-pow}
+  \caption{The user interface showing three browser tabs opened and competing for mining the next blocks, 
+  while achieving consensus and presenting the same view.}
+  \label{fig:in-browser-pow}
+\end{figure}
+
 \subsection{Repository}

### 2022-11-19 practical-consensus/src/Block.ts

@@ -36,6 +36,6 @@ export class Block {
     }
 
     static genesisBlock(): Block {
-        return new Block('0000003466f7289d52851720ad259c88f192604759e34ece83d6eb74fa03b5ed', '', 9865212, 'edurbrito');
+        return new Block('04ff496d58aae22254d59665a4a5e9dd2a011247b2414c5d8b8456a0353f5761', '', 112358, 'dddddd-edurbrito');
     }
 }

### 2022-12-03 theoretical-consensus/article/article.tex
change hyphens
refactor on feedback

@@
 % correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
+\hyphenation{transi-tioning computa-tionally reaching tole-rance parti-cipants extrapo-late determi-nistic approach approaches issues capabi-lities limi-tations commu-nication}
 
@@
 or bottlenecks, but also in terms of scalability \cite{miller2016honey}. Trying to put some effort on decentralization, 
-Byzantine-Fault-Tolerant consensus protocols are not known for their performance when the network grows
+Byzantine-Fault-Tolerant consensus protocols, hereinafter abbreviated as BFT protocols, 
+are not known for their performance when the network grows
 in size. The more participants there are, the more messages need to be exchanged
@@
 There are a couple of particularities that need to be mentioned. First, the block generation
-is of probabilistic nature, as the nonce is a random value, and the hash function is a one-way function.
+is of probabilistic nature, as the nonce is a random value, often of the form of a counter, 
+and the hash function is a one-way function.
 Allied to this, increasing the hashing power is not an easy task, which consequently allows for tackling Sybil attacks,
 despite the permissionless and pseudonymity nature of the network. Plus, the incentive mechanism
-also plays a role in this regard, by naturally incentivizing honest participation.
+also plays a role in this regard, by naturally incentivizing honest participation, at the application level, 
+for example, in the form of monetary rewards, in the case of Bitcoin \cite{nakamoto2008bitcoin}.
 The block generation interval, forged by the adjustment of the difficulty
@@
 consensus protocol, not at the block proposal level, but at the
-block acceptance level. The idea is to have a free flow of blocks
+block validation or acceptance level (See Fig. \ref{fig:building-blocks-consensus}).
+The idea is to have a free flow of blocks
 proposed by the network, but to have a quorum of entities that
@@
 However, there are some key aspects that need to be addressed
-in order to make this approach viable for resource-constrained networks, as we saw in the previous section:
+in order to make this approach viable for resource-constrained networks, with regard to their characteristics. These aspects are:
 
@@
 protocols, and the challenges that arise when applying these protocols to different environments.
-In particular, this paper presented the challenges that resource-constrained networks may be facing, 
+We reasoned about the need for permissionless consensus, and showed that most of the existing
+consensus protocols follow a very similar phased design, often differing only in the block
+generation phase, or in the consensus proof units, concluding that those design choices are usually 
+related with the trade-offs imposed by the blockchain trilemma.
+We also presented some other approaches that deviate from that common
+structural design, and that may be more suitable for some specific network environments.
+In particular, this paper finished presenting the challenges that resource-constrained networks may be facing, 
 and the limitations of the existing consensus protocols when applied to this type of networks.
-We have also presented the main approaches that have been explored to address these challenges,
+We also argued about the main approaches that have been explored to address these challenges,
 and showed some potential paths for future research.
